Reversing protonation of weakly basic drugs greatly enhances intracellular diffusion and decreases lysosomal sequestration

Version 1

Abstract

Abstract

For drugs to be active they have to reach their targets. Within cells this requires free diffusion, distribution, and availability. Here, we explored the intracellular diffusion rates and distribution of a series of small molecular fluorescent drugs, in comparison to proteins, by microscopy and fluorescence recovery after photobleaching (FRAP). While all proteins diffused freely, we found a strong correlation between pKa and the intracellular diffusion and distribution of small molecule drugs. Weakly basic, small-molecule drugs displayed lower fractional recovery after photobleaching and 10-to-20-fold slower diffusion rates in cells than in aqueous solutions. As, more than half of pharmaceutical drugs are weakly basic, they, are protonated at pH 7.4, resulting in their sequestration in the cell cytoplasm. Protonation, facilitates the formation of membrane impermeable ionic form of the weak base small molecules. This results in ion trapping, further reducing diffusion rates of weakly basic small molecule drugs under macromolecular crowding conditions where other nonspecific interactions become more relevant and dominant. Our imaging studies showed that acidic organelles, particularly the lysosome, captured these molecules. Surprisingly, blocking lysosomal import only slightly increased diffusion rates and fractional recovery. Conversely, blocking protonation by N-acetylated analogues, greatly enhanced their diffusion and fractional recovery after FRAP. Based on these results, N-acetylation of small molecule drugs may improve the intracellular availability and distribution of weakly basic, small molecule drugs.

Introduction

Drug molecules alter many biochemical reaction pathways inside the cell by interacting with proteins, DNAs, RNAs, or others. By this, they influence one or more specific cellular activities (Ong et al., 2009; Schreiber, 2005; Stockwell, 2004). Due to their bioactivity, stability, rather simple chemical synthesis allowing for industrial bulk scale production and the easiness of administration, most drugs are small molecules (< 1 kDa) (Ngo and Garneau-Tsodikova, 2018; Veber et al., 2002). Over half of these drugs have a significant degree of hydrophobicity and are weakly basic at intracellular pH of 7.4 (Charifson and Walters, 2014; Lipinski et al., 1997). Hydrophobicity is required to cross the lipid bilayer of the cell membranes, and polar functional groups solubilize the drug in the cellular medium (Lipinski et al., 1997). The balance between pKa and logP (partition coefficient of small molecules as measured from their partition between octanol/water) is critical for successful drug design (Manallack, 2007; Meanwell, 2011). Lipinski’s “Rules of 5” suggests that logP should be between 1 and 6 to be a candidate for successful oral administration. Small molecule drugs can be acidic or basic depending on their pKa values (Lipinski et al., 1997). Drugs with pKa ∼ 0-7 are considered acidic, and pKa ∼ 8-14 are regarded as basic or alkaline. Basic drugs with pKa ˃8 are more common than their acidic counterparts (pKa ˂ 7, ∼60:40 ratios). In some cases, lowering their pKa to the range between 6 and 7, has proven beneficial (Charifson and Walters, 2014). For a long time, mechanisms such as passive diffusion (Di et al., 2012) and carrier-mediated drug delivery (Dobson and Kell, 2008) or a combination of both (Sugano et al., 2010) have been used to explain intracellular drug delivery. However, transport of a drug across the cellular membrane is only one part of a successful delivery. Intracellular diffusion and bio-distribution needs to be taken in account when assessing the drug mode of action.

In previous work, we studied the diffusion of fluorescent drug molecules in crowded environments in vitro (Dey et al., 2022). However, the cellular cytoplasm of eukaryotic cells is by far more complex (Model et al., 2021). Intracellular membranes, multiple organelles (Ellis, 2001; Fulton, 1982) and a large diversity of macromolecules, in terms of size, charge and hydrophobicity, render the cellular interior extremely complex and heterogeneous (Milo and Phillips, 2015; Neurohr and Amon, 2020). Moreover, the pH inside cellular organelles can vastly differ from the cytosolic pH (Madshus, 1988). Organelles like lysosomes, late endosomes, Golgi, and secretory vesicles are acidic and have been known for sequestering weakly basic drugs (Asokan and Cho, 2002; Proksch, 2018). Therefore, predicting diffusion of a particular drug molecule in the dense cytoplasm is a complex task and requires sub-organelle resolution. For example, positively charged (basic) drugs designed to target nuclear DNA may bind cytoplasmic matter through nonspecific interactions (Long et al., 2022; Tiwari et al., 2018), they could be sequestrated within the lysosome, or diffuse slower than expected due to crowding and nonspecific interactions (Gotink et al., 2011; Jansen et al., 1999; Zhitomirsky and Assaraf, 2015).

The mobility of a drug in the cellular interior has been shown to be a good proxy for intracellular drug delivery and other processes, such as nonspecific binding or aggregation (Dey et al., 2022). The primary tool for measuring small molecule diffusion inside the cell is fluorescence recovery after photobleaching (FRAP), which inherently requires fluorescent drug molecules. FRAP with a confocal microscope is a sufficiently sensitive, versatile, and easy-to-use method suitable for measuring the diffusion of weakly fluorescent molecules. Unfortunately, single-molecule fluorescence correlation spectroscopy (FCS), a gold-standard method for measuring diffusion of fluorescent particles, cannot be applied for most small molecule drugs due to their weak fluorescence (Dey et al., 2022, 2021; Zotter et al., 2017).

In this work, we determined the intracellular diffusion coefficients of a series of fluorescent dyes and drugs compared to protein diffusion by FRAP (Figure 1 and Table 1). Specifically, we used Line-FRAP, which has a much higher time resolution and therefore can measure faster diffusion rates in comparison to standard FRAP (800 Hz versus 50 Hz, for detailed explanations, see (Dey et al., 2022) (Dey et al., 2021)). The apparent diffusion coefficients derived from FRAP measurements are termed Dconfocal (Equation 1, materials and method sections). Importantly, Dconfocal values for proteins determined by Line-FRAP agreed with the diffusion rates measured by FCS (Dey et al., 2021). For small molecules, Line-FRAP only provides relative 3D diffusion coefficients (Dey et al., 2022). We studied the following fluorescent drugs: the antimalarial drugs quinacrine dihydrochloride (Ehsanian et al., 2011), primaquine phosphate (Camarda et al., 2019) and amidoquine (Olliaro and Mussano, 2009). The anti-cancer drug mitoxantrone (Shenkenberg and Von Hoff, 1986), and the glycogen synthase kinase GSK3 inhibitor SB216763 (Wagman et al., 2005). In addition, we monitored the diffusion of the diagnostic staining agent fluorescein disodium salt (Jampol and Cunha-Vaz, 1984), the fluorogenic substrate CCF2 (Zuverink and Barbieri, 2015) and boron dipyrromethene (BODIPY) analogues, which are used as fluorescent markers for lipids, membranes, and other lipophilic compounds (Poryvai et al., 2022).

Chemical structures and pKa values (prediction from ChemAxon software) of the small molecules and drugs used in this study.

The diffusion coefficients and their cellular distributions showed very slow diffusion of the basic compounds, i.e., low FRAP recovery, combined with an accumulation of the molecules in lysosomes. Surprisingly, while inhibiting the V-ATPase H+ pump using Bafilomycin A1 (Spugnini et al., 2010; Wang et al., 2021) or sodium azide (Hiruma et al., 2007) inhibited accumulation in lysosomes, cellular diffusion and fractional recovery increased only slightly. Contrary, blocking protonatable amino groups by acetylation greatly enhanced diffusion and FRAP recovery of studied small molecules. As over half the small molecule drugs are basic, our findings are highly important in future drug development, as slow diffusion and low FRAP recovery is directly related to reduced cellular activity. Contrary to small molecules, we show here that diffusion rates within the cytoplasm of HeLa cells of 16 different E.coli proteins with pKa values of 4.5-8, are within the expected range, suggesting that hindered diffusion is mainly a problem of small molecules.

Results

Acidic Proteins diffuse freely in the HeLa cytoplasm

We have previously shown that diffusion in the cellular cytoplasm of E-fts and baeR, two bacterial globular proteins, aligns with the expected values considering the ∼2-3-fold higher macro-viscosity (Dey et al., 2021). Here, we extended this study to include 16 additional proteins (mostly from E. coli) whose oligomeric states in solution were previously characterized (Table S1 and (Marciano et al., 2022)). The idea was to see and compare how a considerable population of proteins, heterologous to the cell (HeLa) diffuse inside the living cell. The proteins showed Dconfocal values of 18-30 µm2s-1 in HeLa cell cytoplasm, which is typical for a cytoplasmic protein diffusion coefficient (Figure 2). Moreover, a complete FRAP recovery was observed, indicating free diffusion. Table S1 provides the predicted isoelectric points and net charges at pH 7.4 of the purified proteins as calculated by Prot Pi, showing pKa values of 4.5-8, as are the majority of E.coli cytosolic proteins. The results show that acidic and neutral proteins, even if not in their native environment (E.coli proteins in HeLa cells), do not stick to the HeLa cell cytoplasm. Next, we studied diffusion of small molecule drugs in HeLa cells.

Diffusion coefficients of proteins and small molecules in the HeLa cells.

Comparison of diffusion coefficients (Dconfocal in blue circles) and percentage of recoveries (in red squares) of bacterial proteins and BSA as measured in HeLa cell cytoplasm are shown. Error bars represent SE calculated from fitting the FRAP progression curves, which are averaged over at least 30 independent measurements.

Negatively charged small molecules rapidly diffuse in live cells

Fluorescein is a negatively charged small organic molecule (pKa’s = 2.85, 3.7, M.W. = 332 Da, Figure 1 and Table 1). We showed in previous work that fluorescein diffusion is slower in the presence of crowder proteins like BSA, lysozyme and, myoglobin, even at relatively low protein concentrations (Dey et al., 2022). Here, the Fluorescein dye was micro-injected in live HeLa cells, followed by FRAP measurements. For comparison, we also incubated HeLa cells for 24 hrs with fluorescein, which is a sufficient time for fluorescein to accumulate in the cells by diffusion. We found that the diffusion coefficient of fluorescein in PBS (Dconfocal of 56.5±2.4 µm2s-1) was reduced to Dconfocal of 38.5±2.0 µm2s-1 in HeLa cytoplasm (Figure 3B). In the cell nucleus, a value of Dconfocal of 43.2±2.2 µm2s-1 was determined, while in concentrated HeLa cell extract (equivalent protein concentration of ∼100 mg/mL), Dconfocal of 49.6±1.9 µm2s-1 was measured. The faster Dconfocal in cell extract compared to live-cell cytoplasm may result from the higher crowding density of the latter. The lowering of diffusion coefficients of negatively charged fluorescein correlates with the increasing crowding density of the medium moving from simple buffer to cells. While micrographs did not show dye aggregation inside the cell cytoplasm (Figure S1A-B) and diffusion rates are the same as determined by microinjection (Figure S1C and S1D), the fractional recovery dropped from 0.95 to 0.65. This suggests that a fraction of fluorescein is immobile in the cell after 24 h incubation.

Fluorescein, CCF2 and CF514 diffusion in PBS and inside HeLa cells.

Comparative averaged FRAP profiles (N= 30; R= 0.99 for each of the fits) in PBS and inside HeLa cells with exponential fits are shown for (A) Fluorescein, (D) CCF2 and (G) CF514. Comparative Dconfocal values for (B) Fluorescein (E) CCF2 and (H) CF514 are also shown. HeLa cells after Micro-injection are shown in (C) for Fluorescein (F) for CCF2 and (I) for CF514. Error bars represent SE calculated from fitting the FRAP progression curves, which are averaged over at least 30 independent measurements.

CCF2 is a fluorogenic substrate used for monitoring β-lactamase activity, with a pKa of ∼5.1 (Zlokarnik et al., 1998). It is composed of cephalosporin core linking B7-hydroxycoumarin to fluorescein (Figure 1). Its calculated Dconfocal values in PBS and inside cells are about half as fast as those measured for fluorescein (Figure 3D-F), and its fractional recovery is ∼0.7. No sign of aggregation is seen in micrographs inside the cell cytoplasm or nucleus (Figure 3F). The substantial slower rate of recovery, and lower fractional recovery of CCF2 relative to Fluorescein suggests its diffusion to be obstructed within the cell. This is in line with our prediction, that the observed substrate limited enzymatic degradation of CCF2 in HeLa cells is a consequence of its occlusion within the cellular cytoplasm (Zotter et al., 2017).

A third molecule for which we determined the diffusion coefficient is the labelling dye CF514. It is used to label proteins and has high quantum yield. As its structure has not been published, we estimated the pKa value from the pH dependent UV-spectrophotometry measurements. The sharp changes of the UV-spectrum at pH∼3 (Figure S1E-F), as well as the redshift confirms a pKa value of ∼ 3. The FRAP curves look similar to fluorescein (Figure 3G-I), with ∼90% recovery after FRAP. Dconfocal in PBS was ∼ 61.1±3.0 µm2s-1, and 26.3±1.9 and 29.2±2.0 µm2s-1 in the cell cytoplasm and nucleus of HeLa cells, respectively (Figure 3H). The drop of Dconfocal is attributed to cellular crowding. The results so far show that acidic small molecules diffuse well inside cells. Next, we focus on basic small molecules with higher pKa values.

High FRAP recovery but slow diffusion measured for the glycogen synthase kinase (GSK3) inhibitor SB216763

The GSK3 inhibitor SB216763 (pKa=7.0, M.W.=371 Da, Figure 1) aggregates when dissolved in PBS buffer solution. However, in our previous study we have seen that aggregation of this particular small molecule drug can be reversed by adding BSA and the extent of de-aggregation depends on the amount of BSA added (Dey et al., 2022). Here, our aim is to compare diffusion behavior of the GSK3 inhibitor in solution phase to that found in live cell cytoplasm. Since the molecule has some inherent solubility issues in PBS buffer, we compared its diffusion also in BSA protein solution and in cell culture media DMEM (Dulbecco’s modified eagle medium). Comparative FRAP profiles and Dconfocal rates of SB216763 in different mediums are shown in Figure 4A-C. The FRAP profiles and relative Dconfocal values show very slow diffusion of SB216763 in HeLa cell cytoplasm of ∼0.61±0.03 µm2s-1 with a fraction recovery of 0.8 after 10s (Figure 4A-C). This is a 40-fold reduction in Dconfocal compared to measurements in 100 mg/mL BSA protein solution, where SB216763 is not aggregating (Dconfocal∼24.3±0.5 µm2s-1 and fractional recovery of ∼0.9 - Figure 4A-C). In cell culture DMEM media (containing 5% serum protein) fractional recovery after FRAP is very low (Figure 4A), indicating that the drug is still aggregating (Dey et al., 2022). Interestingly, the Dconfocal value in HeLa cell extract was higher ∼ 11.3±0.7 µm2s-1 as is its fractional recovery (0.9), showing that HeLa cell extract, like BSA is disaggregating SB216763. Co-localization with Lysotracker-633 (Figure 4D-E) shows that SB216763 colocalizes and is sequestered in acidic chambers of the cell lysosomes. Moreover, SB216763 accumulation into the lysosome increased over time (Figure 4D and E, 45-minute and 2-hour treatment). Varying drug dosages resulted in Dconfocal values of 0.42-0.61 µm2s-1, suggesting that the slow diffusion is not related to SB216763 concentration (Figure S2A-B).

Diffusion of GSK3 inhibitor in PBS and inside HeLa cells.

Comparative (A) averaged FRAP recovery profiles with exponential fits for GSK3 inhibitor (SB216763) (N= 30; R= 0.99 for each of the fits). Dots show prebleach fluorescence, Black line for DMEM, blue for PBS buffer + 50 mg/mL of BSA, green for HeLa cells and orange for HeLa extract. (B) Averaged bleach size profiles with gaussian fits (N= 30; R= 0.99 for each of the fits) and (C) Calculated Dconfocal values. (D-E) Colocalization of GSK3 inhibitor in lysosomal compartments. HeLa cells were treated with Lyso Tracker 633 for 30 mins. GSK3 inhibitor treatment for (D) 45 mins and (E) 2 hours respectively at 10 µM concentrations are also shown. Error bars represent SE calculated from fitting the FRAP progression curves, which are averaged over at least 30 independent measurements.

Despite its high solubility in aqueous solution, Quinacrine diffusion in HeLa cells is slow

Quinacrine DHC (M.W.= 472.9 Da, Figure 1) is a compound highly soluble in PBS (Dey et al., 2022) with a fast Dconfocal rate of 41.9±3.0 µm2s-1 (Figure 5A-C). This rate dropped to Dconfocal∼2.5±0.1 µm2s-1 in HeLa cell cytoplasm, with fraction recovery of ∼0.8, in comparison to one in buffer. In HeLa cell extract a Dconfocal rate of 15.1±0.6 µm2s-1 was determined, which lays between PBS and HeLa cytoplasm. Micrographs of live HeLa cells show that the quinacrine DHC molecules aggregate in the cytoplasm. Co-localization with Lysotracker-633 (Figure 5D-E) show that quinacrine colocalizes and is sequestered in acidic chambers of the cell lysosomes. Moreover, quinacrine aggregation into the lysosome increased over time (Figure 5D and E, 45-minute and 2 hours treatment). Therefore, we measured diffusion instantaneously following micro-injection, and compared it to 2 and 24 hours of incubation (Figure S3A-B). Quinacrine accumulation in the lysosome was observed also immediately after micro-injection, with aggregation increasing over time. Dconfocal of 4.2±0.2 µm2s-1 was calculated from line-FRAP immediately after micro-injection, slowing to 2.2±0.1 µm2s-1 following 2 hours incubations, with fractional recoveries of 0.63 and 0.57 respectively. After 24 hours of incubation the fractional recovery is less than 0.25, and thus the diffusion value is meaningless (Figure S3C and D). Next, we compared FRAP of the cells treated with 2 and 6 µM quinacrine solutions after 2- and 24-hours incubations (Figure S4A-F). The concentration of quinacrine had only a marginal effect on the results. Again, following 24 hours of incubation, most of the quinacrine is in the lysosomes.

Diffusion of Quinacrine in PBS and in HeLa cells.

Comparative (A) averaged FRAP recovery profiles with exponential fits for Quinacrine dihydrochloride (N= 30; R= 0.99 for each of the fits), (B) averaged bleach size profile with gaussian fits (N= 30; R= 0.99 for each of the fits) and (C) diffusion coefficients in PBS buffer, in 50mg/mL of BSA, HeLa cell extract and in live HeLa cells are shown. (D-E) Colocalization of Quinacrine DHC in lysosomal compartments are shown. HeLa cells were treated with Lyso Tracker 633 for 30 mins. Quinacrine DHC treatment was done for (D) 45 mins and (E) 2 hours respectively at 10 µM concentration. Error bars represent SE calculated from fitting the FRAP progression curves, which are averaged over at least 30 independent measurements.

Mitoxantrone, primaquine and amidoquine all aggregate within cells

Mitoxantrone (anti-cancer) and primaquine phosphate (antimalarial) fluorescence with low quantum yields and weak photo-stability. However, due to their slow diffusion, the temporal resolution of standard XY-FRAP was sufficient for FRAP measurements (Figure 6A-H). Figures 6A and 6E show very low fraction recovery for these two drugs, suggesting that these weakly basic drugs are also sequestered in the cellular cytoplasm, as indeed shown in Figures 6B and 6F. Colocalization with LysoTracker red-633 show that Primaquine (S5A-B) and Amidoquine (S5C-D) are sequestered in the lysosomes. A summary of the measured FRAP rates in buffer and in cells are shown in Figure S6. The results shown above for the 5 basic drugs led us to investigate whether that their trapping in lysosomes is the reason for their slow diffusion.

Mitoxantrone and Primaquine diffusion inside HeLa cells.

Averaged XY-FRAP recovery profiles with exponential fits (N= 20; R= 0.98 for each of the fits) for (A) Mitoxantrone and (E) Primaquine in HeLa cells. Fluorescent and Transmission channel images of HeLa cells after 45 mins incubations with (B-C) Mitoxantrone and (F-G) Primaquine are shown. Time lapse micrographs of a portion of HeLa cell going through a classical rectangular XY-FRAP protocol for (D) Mitoxatrone and, (H) Primaquine are also shown.

Inhibition of lysosomal sequestration is only slightly increasing diffusion in cells

Lysosome internal pH of ∼4.5 is maintained by vacuolar proton ATPase (V-H+-ATPase) (de Duve et al., 1974; Kaufmann and Krise, 2007). The low pKa of weakly basic small molecule drugs drive their accumulation in the lysosome. There, they become protonated (ionic form is less-permeable), which hinders their back-diffusion to the cytosol. The phenomenon is known as cationic ion trapping (Asokan and Cho, 2002; Kaufmann and Krise, 2007) and results in drug accumulation in lysosomes up to millimolar concentrations (de Duve et al., 1974). We blocked lysosomal acidification using either Bafilomycin A1 or sodium azide prior to drug administration. Bafilomycin A1 is a specific V-ATPase inhibitor and thus inhibits the supply of protons to the lysosomal chamber (Wang et al., 2021). Sodium azide, blocks ATP generation, resulting in the disruption of cellular activities which require ATP (Hiruma et al., 2007; Ishii et al., 2014). As a result, cationic ion trapping is inhibited. These two compounds were used to investigate whether lysosomal trapping is the cause for the 20-40-fold lower diffusion and lower fraction recovery in HeLa cells compared to buffer. Figures S7A and B, and Figure S8A and B show colocalization images for GSK3 inhibitor and quinacrine, respectively (with Lysotracker-633 used as marker) after addition of 100 nM Bafilomycin A1 (Yoshimori et al., 1991). The micrographs show that Bafilomycin A1 strongly reduces lysosomal volumes following quinacrine and GSK3 treatment (Figure S7A vs S7B and S8A vs S8B). However, Dconfocal values were not altered by Bafilomycin A1, as shown in Figures S7C-D and S8C-D for GSK3 inhibitor and Quinacrine, respectively. Dconfocal for the GSK3 inhibitor is ∼ 0.60±0.02 µm2s-1, and for Quinacrine ∼ 2.4±0.1 µm2s-1. This suggests that Lysosomal accumulation of these two drugs is not the reason for the slow diffusion measured in the HeLa cell cytoplasm.

To verify the Bafilomycin results, HeLa cells were pre-treated with Sodium azide at 1 and 50 mM concentrations (Figure 7A and D), followed by treatment with GSK3 inhibitor or Quinacrine (Poste and Papahadjopoulos, 1976). The micrographs in Figures 7A and 7D, show that sodium azide inhibited lysosomal accumulation of these drugs. Comparative FRAP profiles and diffusion coefficients (Figure 7B-C and 7E-F) were not much changed for the two drugs and stayed low (Dconfocal are 1.8±0.1-1.9±0.1 µm2s-1 for quinacrine and 0.4-0.5 µm2s-1 for GSK3 inhibitor), similarly to what was observed with Bafilomycin.

Effect of Sodium Azide on the diffusion of GSK3 inhibitor and Quinacrine inside HeLa cells.

(A) GSK3 inhibitor and (D) Quinacrine dihydrochloride treated HeLa cells with or without sodium azide. Comparison of averaged FRAP profiles with exponential fits (N= 30; R= 0.99 for each of the fits) and calculated Dconfocal values for (B-C) GSK3 inhibitor and (E-F) Quinacrine dihydrochloride. Error bars represent SE calculated from fitting the FRAP progression curves, which are averaged over at least 30 independent measurements.

5-amino fluorescein diffuses slowly in HeLa cells and is accumulated within lysosomes

To experimentally test the relation between the pKa of small molecules and their in cell diffusion, we measured the diffusion of 5-amino fluorescein (AM-Fluorescein, M.W.= 347.3 Da, Figure 8A), which is structurally similar to fluorescein disodium salt, but with an additional-NH2 group that increases its pKa from ∼3 to ∼9 (Figure 8A). Figures 8B and C show comparative FRAP traces for these molecules, and for CCF2. While fluorescein disodium salt has a fast FRAP rate, and is completely recovering, the cationic charged AM-fluorescein recovers much slower with a fractional recovery of 0.4. Moreover, while fluorescein disodium salt is not sequestered in lysosomes, 5-Amino fluorescein is. This sequestration is alleviated by adding sodium azide to the cells (Figure 8D), but has only a small effect on the FRAP rate and fractional recovery (Figure 8B). This experiment strongly suggests that the functional primary amine group is the main contributor towards the slow diffusion and low FRAP recovery of AM-fluorescein in the cell cytoplasm.

FRAP of Fluorescein analogues inside HeLa cells.

(A) Structures of Fluorescein analogues with pKa values. (B) FRAP recoveries with exponential fits for Fluorescein, CCF2 and 5-amino Fluorescein, with/without Na-Azide treatments in HeLa cells (N= 30; R= 0.99 for each of the fits). (C) Line FRAP profiles with time lapses. (D) Micrograph images of treated HeLa cells after small molecule incubations.

Acidification of primaquine and BODIPY increases their diffusion and FRAP recovery

We substituted the NH2 of primaquine (antimalarial drug) and BODIPY compounds for the charge-neutral-NHAc functional group (Figure S9A and S10A). Due to their low quantum yield, we could not implement the faster Line-FRAP protocol, therefore, we used standard FRAP, which suits the slow diffusion of both compounds. As seen from the FRAP profiles (time-dependent monographs) in Figures S9B-D for primaquine derivatives and in Figures S10B-C for BODIPY derivatives, the recovery is much slower for free amine-containing moieties. Figure S9I-J shows that the fractional recovery for primaquine is only ∼0.27, whereas, for primaquine-NHAc, the recovery is ∼0.6. Comparing the measurements using micro-injecting versus incubation of the cells for FRAP measurements for primaquine-NHAc gave similar results (Figure S9E-H). Unfortunately, the classical FRAP is not fast enough to calculate the diffusion coefficients, as the dead time was 150 ms. For the BODIPY analogues, the observations were similar, but to a smaller extend. For BODIPY-NH2, the recovery percentage is ∼57%, whereas, for BODIPY-NHAc, the recovery is ∼67% (Figure S10H). The estimated recovery half-life also supports the faster diffusion of BODIPY-NHAc, as shown in Figure S10I.

Basic small molecules do not colocalize with lipid droplets, the ER or nucleic acids

To identify the location of colocalization of basic small molecules, we tested colocalization of GSK3 inhibitor in lipid droplets using the Nile red dye (Figure S11) and in the ER using a specific BFP/mCherry tagged fluorescent antibody markers (Figure S12). Next, we looked for colocalization of Mitoxantrone with nucleic acids present in cytoplasm using SYTO blue (Figure S13). No colocalization was found between the drugs and either lipid droplets, the ER or nucleic acids. Super resolution images of GSK3 inhibitor treated HeLa cells were taken to improve the spatial distribution (Figure S14), however, this did not contribute new information to the puzzle. Therefore, we could not pinpoint the location of sequestration of weakly basic small molecule drugs within the cellular cytoplasm, except the lysosome.

Discussion

In this study, we selected well-known and widely used small molecule drugs to study their behavior in aqueous solutions and living cells. All drug molecules used here are fluorescent, allowing for their FRAP measurements. Previously, we found that small-molecule drugs diffuse differently in crowded media than in a simple buffer solution, even if aggregation is not an issue (Dey et al., 2022). Here, we show that many small molecules get trapped within the cellular cytoplasm, resulting in extremely slow diffusion and lower fractional recovery after FRAP. This is particularly observed for cationic-charged small molecules, with Dconfocal values of 20-40-fold lower than observed for fluorescein or CF514, which, in many cases is accompanied by low recovery after photobleaching. The later suggests that most molecules are occluded to components of the HeLa cell cytoplasm. Even dense cell extract from HeLa cells is not mimicking the in vivo results (Figure 4A and C). This is very different from protein diffusion, where none of the 16 proteins measured here showed slow diffusion or low fractional recovery, despite their E.coli origin, expressed in HeLa cells. Slow diffusion and lower fractional recovery were limited to basic small molecules, showing Dconfocal values of 0.2-2 µm2s-1. While the fractional recovery for all proteins is ∼0.9, for many basic small molecules it is 0.2-0.5.

As primary culprit for the slow diffusion of the basic small molecule drugs we suspected sequestration in the lysosomes. Indeed, high level of accumulation in the lysosomes were found for all the basic drugs. However, pre-treatment with sodium azide or Bafilomycin A1, which we showed to inhibit lysosomal accumulation, had only a small effect on diffusion or fractional recovery after FRAP, which stayed low. Therefore, we evaluated colocalization of GSK3 inhibitor with lipid droplets, the ER or nucleic acids present in cytoplasm (Figure S13), but did not detect any.

The conclusive evidence that slow diffusion and lower FRAP recovery results from the drugs being basic came by altering three molecules from being basic to acidic and vice versa. The best example here is fluorescein, where transformation to 5-amino fluorescein drastically reduced its diffusion as well as its fractional recovery after FRAP. Two additional examples here were Primaquine that was modified to Primaquine Acetate and BODIPY-NH2 that was modified to BODIPY-NHAc. In both cases, the diffusion became faster and the fractional recovery increased significantly. This finding is important, as the activity of a drug within the cell is dictated by its active concentration. When the drug is sequestered, its active concentration is reduced. Even, if it is not fully sequestered, but only diffuses much slower, it will result in lower association rate constant, and thus lower affinity towards the drug target. As a result, to keep the drug active, its dose has to be increased, which can have negative implications on side effects, due to off-target binding.

Conclusions

Our study shows that in vitro biophysical crowding studies for small molecule drugs are of limited value if one wants to understand the biophysical behavior of the same drug within the cell. While it is true that availability of a drug within a cell is only one factor dictating its biological activity, it is a crucial one. Our findings here raise an important limitation on the standard rules for drug design, as these do not consider the stickiness of basic small molecules within the cell. We were able to directly address this question by using fluorescent molecule drugs, and measuring their location and diffusion within cells. Most importantly, we also show that by blocking protonation of a number of basic compounds we were able to increase substantially their diffusion and recovery after photobleaching, and thus potentially their activity. These findings may be consequential in future drug development.

Materials and methods

All the reagents used are described in a table format (Table S2, Key resources). The supplementary method sections describe the synthetic procedures, characterizations, and purity of the synthesized compounds.

Mammalian Cell Culture

HeLa cells were grown in 35-mm glass-bottomed dishes (MatTek Corporation) in DMEM (1X) Gibco (Life Technologies Limited) supplemented with 1X pyruvate, penicillin/streptomycin (BioIndustries), and 10% fetal bovine serum (Life Technologies Limited). The cells were subcultured when 80% confluence was reached using trypsin-EDTA for cell detachment. 2×105 HeLa cells in 2.5 ml of DMEM were pipetted into glass-bottomed dishes and incubated overnight. The cells were cultured in a humid atmosphere at 37 °C and 5% CO2. The cells were imaged 24–30 h after seeding. Before the microinjection, the medium was aspirated, and fresh medium was supplemented with 25 mM HEPES, pH 7.4. For drug treatment, HeLa cells were incubated with 10 µM of drug (diluted 1:1000 from stock) for indicated times at 37 °C, followed by three times PBS 1X wash prior to imaging. For dose-dependent studies, the drug concentrations varied typically between 2-12 µM. Sodium azide concentration was 1 and 50 µM. Bafilomycin A1 was used at 100 nM concentration. Similarly, during the colocalization study, HeLa cells were treated with LysoTracker dye (1X) with 1:1000 dilution for 30 mins to 1 h at 37 °C, followed by 3x washing with PBS. Quinacrine, GSK3 inhibitor, Mitoxantrone, 5-amino Fluorescein, Amidoquine, Primaquine and BODIPY analogues were incubated with cells, while Fluorescein, Quinacrine, Primaquine-Ac, and CF514 dye were also micro-injected inside the HeLa cells.

Microinjection into HeLa cells

Microinjections were performed using the Eppendorf FemtoJet microinjector attached to the Eppendorf InjectMan NI2 micromanipulator. The fluorescein sodium salt, CCF2, CF514, quinacrine, primaquine analogues, BODIPY analogues were dissolved in DMSO and highly concentrated stock aliquots were made. Diluted PBS solutions of small molecules were injected into cells using glass capillaries from Warner instruments and pulled by a vertical puller (Narishige). For every measurement, a single pressure pulse was applied to deliver the sample into the cell. Air was administrated at 15– 25 hPa for 0.1-0.3 s. For injections, single cells containing morphologically healthy and well-connected HeLa cells were selected. Before and after the microinjection, cell morphology and membrane integrity were confirmed by visually inspecting the injected cells. HeLa cell extract preparation. Cytoplasmic HeLa cell extracts were prepared as described previously with slight modifications (Zotter et al., 2017). HeLa cells from 4×10 cm plates at 80% confluency were washed, treated with trypsin and collected. Pellets were mixed with 500µl of RIPA buffer, IMP40, 5 µL of protease inhibitor and stored in ice. After 20 min, the mixture was centrifuged at 13k×g. This process was repeated twice to obtain concentrated extract. A BSA calibration curve determined the total protein concentration; the final concentration was up to 100 mg/mL. After liquid nitrogen freezing, the final cell extract solution was stored at −80° C.

Protein purification and dye labeling

Bacterial proteins used in this study were purified as described by us previously (Marciano et al., 2022). The dye labelling procedure with CF514 for purified proteins and removal of excess dye was described in detail (Dey et al., 2021).

Confocal microscopy and FRAP Analysis

Confocal microscopy. Images were collected with an Olympus IX81 FluoView FV1000 Spectral/SIM Scanner confocal laser-scanning microscope, using 60X DIC oil-immersion objective, N.A. 1.35. For fluorescein sodium salt fluorescence measurements, excitation was done at 440 nm, using a diode laser at an output power of 1-4 % of maximal intensity for high to low concentrations. In contrast, emission was recorded from 520 to 550 nm using the spectral detection system. For CCF2-FA, excitation was done with the laser at 440 nm using 1-2 % of the maximal intensity, while emission was collected from 470-560 nm with SDM560 emission dichromator as cut-off filter. For quinacrine DHC and GSK3 inhibitor, excitation was done at 488 nm laser using 1-2% of the maximal intensity, while emission was collected from 502-560 nm with SDM560 emission dichromator as cut-off filter. During colocalization experiments of quinacrine DHC and GSK3 inhibitor with LysoTracker-633 dye, the same setting was used for detecting the small molecule drugs in Channel 1 (green). LysoTracker-633 dye was detected in Channel 2 (red) by exciting at 635 nm laser using 1-2 % of the maximal intensity, while emission was collected from 655-755 nm. For mitoxantrone and BODIPY analogues, excitation was done at 635 nm laser using 1-2 % of the maximal power, and emission was collected from 655-755 nm in Channel 2 (red). In the blue channel, excitation was done by 400 nm laser using 2-4% of maximal power, whereas the emission was collected from 420-480 nm with SDM480 emission dichromator as cut-off filter. The same setting is used in the blue channel for the SYTO blue maker (to visualize free nucleic acids) and blue fluorescent protein marker (BFP to visualize the ER). For the mcherry protein marker (to visualize the ER), the same setting for the red channel (Channel 2) is used. For every colocalization experiments, proper blank experiments were performed separately in all the respective channels with the same microscope settings to avoid any bleeding/leakage of emission signals from one Channel to another Channel, as shown in Figure S15. Image analyses were performed using FluoView/Imaris software, and data analyses were performed using Kaleidagraph software version 4.1 (Synergy). Colocalization coefficients as a percentage of material colocalized are calculated using Imaris software version 7.

Line-FRAP and Classical XY-FRAP

Line-FRAP was carried out in liquid drops and HeLa cells. For photobleaching, “Tornado” of 4 pixels (2×2) diameter (Dey et al., 2021). It is the smallest area achievable using Tornado. The bleach circle was kept precisely in the middle of the scanning line. The lines we scanned were unidirectional with time intervals of 1.256 ms 1000 times (equivalent to 1.256 s) in the majority of the measurements. The number of scans prior to, during, and after the photobleaching was 10, 42, and 948, respectively. Only in some specific cellular measurements (for GSK3 inhibitor and Quinacrine DHC, where recovery is very slow) the unidirectional lines were scanned 5000 times. Photobleaching was achieved by using the laser at 405-nm or with 635-nm excitations for the 63-millisecond duration at full intensity (100%). The simultaneous scanner moved at a 100 µs/pixel speed to perform an efficient photo-bleach. We have used two simultaneous scanners during the Line-FRAP experiments: one scanner (at 405 nm with the full intensity of 100%) for photobleaching and another scanner (at 440/515/635 nm with weak intensity) for data acquisition.

For all the drugs except Mitoxantrone and BODIPY analogues (emission in red wavelengths), photo-bleach was performed by 405 nm laser. For fluorescence signal detection: fluorescein 440 nm laser (1-4%); CCF2 440 nm laser (1-2%); GSK3 inhibitor 440 nm laser (0-2%); quinacrine (1-2%) of maximal intensity were used. Emission collections were done from 520-550 nm for quinacrine DHC and GSK3 inhibitor.

Line FRAP was not possible for primaquine analogues, as it suffers from strong photo-bleach during recovery. Here classical XY FRAP was performed, with photo-bleach done by 405 nm laser with full intensity. The frame rate is maintained at 172 ms with a bleach pulse duration was 150 ms. Due to microscope setting limitations, Line FRAP is also not feasible for Mitoxantrone and BODIPY analogues. A classical XY-FRAP using two simultaneous scanners is done for these molecules. Here, bleach was performed by simultaneous 635 nm laser with the full intensity of 100%, and the main excitation was done by 635 nm laser (with the power of 4%). Emission collections were done from 655-755 nm. For BODIPY analogues FRAP recoveries, the bleach pulse time was 150 ms. The bleach pulse time was also 150 ms for primaquine analogues for comparative FRAP profile collections. The frame rate is maintained at 172 ms. Using the Olympus IX81 FluoView FV1000 Spectral/SIM Scanner confocal laser-scanning microscope, using Tornado (which requires SIM scanner to be loaded) greatly enhances bleaching efficiency. In addition, it shortens the time to obtain the first measurement after bleach (which is immediate in this mode). This property is highly beneficial for Line-FRAP measurements, where the time scale of data acquisition plays an important role. The Fluoview SIM scanner unit synchronizes laser light simulation with confocal and multiphoton imaging to avoid interruption to image observation during laser stimulation or manipulation. We have varied the intensity of the lasers to achieve an optimal signal/noise ratio. Fluorescence recovery plots were fitted to a double exponent growth curve. FRAP experiments were also performed inside the PBS buffer drops and crowding conditions. A coverslip was introduced at the top of the drop to stop evaporation during measurements. Our previously developed method on Line FRAP protocol has been employed to calculate diffusion coefficients from the FRAP rates and averaged bleach sizes. Diffusion rates cannot be calculated using XY-FRAP. However, FRAP parameters were kept constant so that qualitatively we could compare the average half-life and percentage of recoveries.

Determination of Dconfocal from FRAP results

We previously developed (Dey et al., 2021) the FRAP in Line mode (Line FRAP) method to monitor the diffusion rates of proteins in various environments. Line-FRAP allows a much faster data acquisition rate compared to classical XY FRAP. It is more suitable for measuring diffusion rates for fast diffusing molecules. The apparent diffusion coefficients calculated from Line FRAP measurements are Dconfocal, which are calculated according to equation 1: where t1/2 is the half-time of recovery and re and rn is the effective and nominal bleach radii values (for details see (Dey et al., 2021)).

To calculate Dconfocal at least 30 independent measurements on different cells (n) were binned, and the curve fit of the progression curve was used to obtain t1/2, and re values and their associated errors as determined from the fit of the curve. rn was obtained from bleaching a fixed sample. The standard errors (SE) of the individual parameters were combined to obtain Dconfocal±SE. To verify SE values, we repeated the 30 measurements independently multiple times, which gave the same SE as obtained from individual curve fits of n = 30 cells.

eLife Assessment

This is a valuable study on the diffusion rates of drug molecules in human-derived cells, highlighting that their diffusion behavior depends on their charged state. It proposes that blocking drug protonation enhances diffusion and fractional recovery, suggesting improved intracellular availability of weakly basic drugs. The correlation between pKa and intracellular diffusion is solid and well-supported, but the study would benefit from a more rigorous statistical treatment and a balanced comparison across different types of compounds. Despite these limitations, the findings are significant for drug design and understanding the biophysical behavior of small molecules in cells.

Reviewer #1 (Public Review):

Summary:

The authors set out to measure the diffusion of small drug molecules inside live cells. To do this, they selected a range of flourescent drugs, as well as some commonly used dyes, and used FRAP to quantify their diffusion. The authors find that drugs diffuse and localize within the cell in a way that is weakly correalted with their charge, with positively charged molecules displaying dramatically slower diffusion and a high degree of subcellular localization.

The study is important because it points at an important issue related to the way drugs behave inside cells beyond the simple "IC50" metric (a decidedly mesoscopic/systemic value). The authors conclude, and I agree, that their results point to nuanced effects that are governed by drug chemistry that could be optimized to make them more effective.

Strengths:

The work examines an understudied aspect of drug delivery.

The work uses well-established methodologies to measure diffusion in cells

The work provides an extensive dataset, covering a range of chemistries that are common in small molecule drug design

The authors consider several explanations as to the origin of changes in cellular diffusion

Weaknesses:

The results are described qualitatively, despite quantitative data that can be used to infer the strength of the proposed correlations.

The statistical treatment of the data is not rigorous and not visualized according to best practices, making it difficult for readers to assess the significance of the findings.

Some important aspects of drug behavior are not discussed quantitatively, such as the cell-to-cell or subcellular variability in concentration.

It is unclear if the observed behavior of each drug in the cell actually relates to its efficacy - though this is clearly beyond the scope of this specific work.

Reviewer #2 (Public Review):

Summary: Blocking a weak base compound's protonation increased intracellular diffusion and fractional recovery in the cytoplasm, which may improve the intracellular availability and distribution of weakly basic, small molecule drugs and be impactful in future drug development.

Strengths:

1. The intracellular distribution of drugs and the chemical properties that drive their distribution are much needed in the literature. Thus, the idea behind this paper is of relevance.

2. The study used common compounds that were relevant to others.

3. Altering a compound's pKa value and measuring cytosolic diffusion rates certainly is inciteful on how weak base drugs and their relatively high pKa values affect distribution and pharmacokinetics. This particular experiment demonstrated relevance to drug targeting and drug development.

4. The manuscript was fairly well written.

Weaknesses:

1. Small sample sizes. 2 acids and 1 neutral compound vs 6 weak bases (Figure 1).

2. A comparison between the percentage of neutral and weak base drug accumulation in lysosomes would have helped indicate weak base ion trapping. Such a comparison would have strengthened this study.

3. When cytosolic diffusion rates of compounds were measured, were the lysosomes extracted from the image using Imaris to determine a realistic cytosolic value? In real-time, lysosomes move through the cytosol at different rates. Because weak base drugs get trapped, it is likely the movement of a weak base in the lysosome being measured rather than the movement of a weak base itself throughout the cytosol. This was unclear in the methods. Please explain.

4. Because weak base drugs can be protonated in the cytoplasm, the authors need to elaborate on why they thought that inhibiting lysosome accumulation of weak bases would increase cytosolic diffusion rates. Ion trapping is different than "micrometers per second" in the cytosol. Moreover, treating cells with sodium azide de-acidifies lysosomes and acidifies the cytosol; thus, more protons in the cytosol means more protonation of weak base drugs. The diffusion rates were slowed down in the presence of lysosome inhibition (Figure 7), which is more fitting of the story about blocking protonation increases diffusion rates, but in this case, increasing cytosolic protonation via lysosome de-acidification agents decreases diffusion rates. Please elaborate.

A discussion of the likely impact:

The manuscript certainly adds another dimension to the field of intracellular drug distribution, but the manuscript needs to be strengthened in its current form. Additional experiments need to be included, and there are clarifications in the manuscript that need to be addressed. Once these issues are resolved, then the manuscript, if the conclusions are further strengthened, is much needed and would be inciteful to drug development.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public Review):

Summary:

The authors set out to measure the diffusion of small drug molecules inside live cells. To do this, they selected a range of flourescent drugs, as well as some commonly used dyes, and used FRAP to quantify their diffusion. The authors find that drugs diffuse and localize within the cell in a way that is weakly correalted with their charge, with positively charged molecules displaying dramatically slower diffusion and a high degree of subcellular localization.

The study is important because it points at an important issue related to the way drugs behave inside cells beyond the simple "IC50" metric (a decidedly mesoscopic/systemic value). The authors conclude, and I agree, that their results point to nuanced effects that are governed by drug chemistry that could be optimized to make them more effective.

We are grateful to the reviewer for summarizing the work and appreciate him/her pointing out that it is high time to consider the drug aggregation and high degree of subcellular localization while optimizing to make them more effective beyond the mesoscopic value like "IC50".

Strengths:

The work examines an understudied aspect of drug delivery.

The work uses well-established methodologies to measure diffusion in cells

The work provides an extensive dataset, covering a range of chemistries that are common in small molecule drug design

The authors consider several explanations as to the origin of changes in cellular diffusion

We are grateful to the reviewer for pointing out the strengths of the manuscript.

Weaknesses:

The results are described qualitatively, despite quantitative data that can be used to infer the strength of the proposed correlations.

The statistical treatment of the data is not rigorous and not visualized according to best practices, making it difficult for readers to assess the significance of the findings.

Some important aspects of drug behavior are not discussed quantitatively, such as the cell-to-cell or subcellular variability in concentration.

It is unclear if the observed behavior of each drug in the cell actually relates to its efficacy - though this is clearly beyond the scope of this specific work.

We have addressed the weaknesses found by the reviewer (see bellow in Reviewer #1 Recommendations For The Authors). Concerning the last point, It would have been indeed very valuable to find a relation between drug's observable behavior and their efficacy, but as the reviewer indicates, it is beyond the scope of this work.

Reviewer #2 (Public Review):

Summary:

Blocking a weak base compound's protonation increased intracellular diffusion and fractional recovery in the cytoplasm, which may improve the intracellular availability and distribution of weakly basic, small molecule drugs and be impactful in future drug development.

We are thankful to the reviewer for summarizing our work and acknowledging that the points raised above can be impactful in future drug development.

Strengths:

(1) The intracellular distribution of drugs and the chemical properties that drive their distribution are much needed in the literature. Thus, the idea behind this paper is of relevance.

(2) The study used common compounds that were relevant to others.

(3) Altering a compound's pKa value and measuring cytosolic diffusion rates certainly is inciteful on how weak base drugs and their relatively high pKa values affect distribution and pharmacokinetics. This particular experiment demonstrated relevance to drug targeting and drug development.

(4) The manuscript was fairly well written.

We are thankful to the reviewer for pointing out the strengths of the manuscript like the intracellular distribution of drugs and properties that drive it, which are missing in the literature.

Weaknesses:

(1) Small sample sizes. 2 acids and 1 neutral compound vs 6 weak bases (Figure 1).

We fully agree with the reviewer on this point. However, the major limitation we have faced here is the small number of drug/drug-like molecules that fluorescent with sufficient high quantum yields. For this study, we initially screened 1600 drugs for their fluorescence in the visible spectrum, and penetration into cells, resulting in 16 drugs. Of those, a small number was suitable for FRAP due to low quantum yield. For some of the molecules (Mitoxantrone, Priaquine), recovery was minimal, making them challenging to study. We added this information in the materials and method section under “Selection of drugs used in this study” (p.10).

(2) A comparison between the percentage of neutral and weak base drug accumulation in lysosomes would have helped indicate weak base ion trapping. Such a comparison would have strengthened this study.

For weakly basic compounds, the ionic form and the non-ionic form of the molecules always remain in equilibrium. The direction of the equilibrium depends on the pH of the medium, which determines the major form of the drug molecules in the solution. Our examples of GSK3 inhibitor (neutral compound, pka~7.0, as predicted by Chemaxon), shows behaviour very similar to the other basic drugs (pka>8) inside the cells. As lysosome pH is about 5.0, the neutral drug also gets protonated inside the lysosomes, as the colocalization study reveals (Figure 4). We added Fig S16 C-D, where we show co-localization of three drugs within the lysosomes showing that all the three weak base drugs colocalize to acidic lysosomes from moderately to extensively. See also in p. 11 under “Confocal microscopy and FRAP Analysis section”.

(3) When cytosolic diffusion rates of compounds were measured, were the lysosomes extracted from the image using Imaris to determine a realistic cytosolic value? In real-time, lysosomes move through the cytosol at different rates. Because weak base drugs get trapped, it is likely the movement of a weak base in the lysosome being measured rather than the movement of a weak base itself throughout the cytosol. This was unclear in the methods. Please explain.

We want to thank the reviewer for pointing this out. To clarify the point, we added to the material and method section in p. 13 the following text: “When the areas of bleach were selected in the drug-treated cell cytoplasm, we avoided the lysosomes as much as possible, within the resolution limits of the confocal microscope. Lysosomes themselves were measured to move within the cytoplasm with an diffusion coefficient of 0.03-0.071 µm2 s−1 (Bandyopadhyay et al., 2014), which is much slower than the diffusion measured for even the slowest compounds using fast Line FRAP, further validating that we did not measure lysosome diffusion.” In addition, we show that in cells after Bafilomycin A1 or Na-Azide treatments the number of lysosomes was reduced drastically (Figures S8& S9, and Figure 7), while the rates of diffusion remain very slow, similar to those measured without lysosomal inhibitors.

(4) Because weak base drugs can be protonated in the cytoplasm, the authors need to elaborate on why they thought that inhibiting lysosome accumulation of weak bases would increase cytosolic diffusion rates. Ion trapping is different than "micrometers per second" in the cytosol. Moreover, treating cells with sodium azide de-acidifies lysosomes and acidifies the cytosol; thus, more protons in the cytosol means more protonation of weak base drugs. The diffusion rates were slowed down in the presence of lysosome inhibition (Figure 7), which is more fitting of the story about blocking protonation increases diffusion rates, but in this case, increasing cytosolic protonation via lysosome de-acidification agents decreases diffusion rates. Please elaborate.

We thank the reviewer for the comment. We added to the results in p. 7 (top) the following “While we selected bleach spots to be small and located outside of lysosomes, this does not assure that some of the bleached area does not include smaller lysosomes. Therefore we investigated whether inhibiting lysosomal trapping will eliminate slow diffusion of cationic drugs.” In addition, we added to the results in p. 7-8 the following: “Comparative FRAP profiles and diffusion coefficients (Figure 7B-D and 7F-H) were slow, but conversely to Bafilomycin, sodium azide treatment did cause a further reduction is rates from Dconfocal 2.4±0.1 µm2s-1 to 1.8±0.1µm2s-1 for quinacrine and from 0.6 to 0.45 µm2s-1 for the GSK3 inhibitor (Figure 7C and G). Both Bafilomycin and sodium azide treatments resulted in elimination of drug confinement in the lysosome, and the small difference in diffusion rates may be a result of the de-acidification of the lysosomes by sodium azide, which may increase the protons in the cytosol upon treatment.”

Reviewer : A discussion of the likely impact:

The manuscript certainly adds another dimension to the field of intracellular drug distribution, but the manuscript needs to be strengthened in its current form. Additional experiments need to be included, and there are clarifications in the manuscript that need to be addressed. Once these issues are resolved, then the manuscript, if the conclusions are further strengthened, is much needed and would be inciteful to drug development.

Reviewer #1 (Recommendations For The Authors):

Major issues:

The paper suffers from poor statistical treatment of the data. FRAP recovery curves should be shown for each repeat, overlaid by an average with SDs as errorbars or shaded regions shown. In bar plots, SEMs should be eliminated in favor of StdDevs. All datapoints should be shown for each bar in Figs. 3-8. To show differences in D_confocal appropriate statistical tests should be conducted. In addition it is unclear what an "independent repeat" is. Does this mean 30 separate imaging sessions/drug treatments/etc? Is it 30 cells on the same coverslip? Is it a combination of both? All reported errors, SD or SEM, should have a single significant digit. Guidelines and best practices for representing quantitative imaging data are all described and visualized in detail in Lord et al. JBS 2020.

We improved the statistics and added the individual progression curves and did the statistics on them as requested. See Figure S2 for individual FRAP curves of fluorescein, GSK3 inhibitor and and quinacrine. Statistical analysis of the individual FRAP curves is in Figure 3B, 4B, 5B, 7C and G. For details see figures legends and material and methods p. 13 in “Determination of Dconfocal from FRAP results”. Line FRAP was done from the cells taken from different plates, treated independently (see text p. 13).

The extensive (and commendable!) dataset the authors have collected can be put to better use than what is currently done. The main text figures in the current form of the preprint are mostly descriptive and their discussion is qualitative, to the point where the author's conclusions are supported only anecdotally. Instead, I would much rather see panels that collate the entire dataset (both protein and drugs) numerically, comparing diffusion values in buffer/cytoplasm/nucleus for all drugs (Like Fig. S6, which is in my opinion the most important in the paper but for some reason relegated to the SI). In addition I would like to see correlations within the dataset, such as D_confocal vs. pKa, vs. concentration (as measured by overall fluorescence signal, see my comment below), vs. mw, or vs. specific chemical moieties (number of charges, aromatic rings, etc). Such correlations should be discussed in terms of a correlation coefficient if conclusions were to be drawn from them, and include errors if available.

We want to thank the reviewer for these suggestions. We now made new Figures 9, and S16 to compare multiple parameters. Figure 9C shows a clear relation between pKa and Dconfocal, but no relation was found between logP, MW or number of aromatic rings and Dconfocal. Fig. S3 also shows the relation between drug concentration and Dconfocal values. These data are now discussed in the discussion section in p. 9 (bottom).

The drug sequestration hypothesis and other conclusions brought forth by the authors could be further tested by looking at the concentration dependence of the drugs inside eachcell and/or its partitioning between different subcellular compartments. The concentration dependence of these drugs is discussed in a very anecdotal fashion using two concentrations - and despite some cases showing an effect no further studies were done. Drug concentrations in this experiment can vary between cells between repeats or even within a single repeat as a result of drug chemistry and delivery methods (microinjection/passive permeability). This is especially important since it is unclear what clinically-relevant concentrations are for each drug (or at least an IC50 for the cell types tested here). I would like to see a quantitative measure of concentrations as another metric to compare diffusion behavior (see my comment above as well).

And maybe one thing to consider in addition would be some discussion in the paper about what sub-cellular distributions might actually mean in the context of drug efficacy (asking for myself as well!) - a paragraph describing recent works on the topic with some references could be instructive.

We want to thank the reviewer for the suggestion. We added now Figure S3, showing the relation between fluorescence intensity in each cell (which is directly related to the concentration of the compound) and FRAP rates and percent recovery for fluorescein, GSK inhibitor and Quinacrine. The results show now relation between drug concentration and FRAP rates, and some relation towards percent recovery. These data are now discussed in the main text (p. 4 bottor and p.6) and in the discussion (p. 9, bottom).

Minor issues:

Readers could benefit from a schematic showing the line FRAP method. It is difficult to understand from the text.

We show now in Figure 2 the line-FRAP method, and discuss it in the introduction (p. 3 top).

Have the authors considered enrichment in the cell membrane? Summed intensity projections or co-labeling with membrane dyes could prove useful to identify if the membrane is enriched in fluorescence.

The microscopy slides, including the super-resolution image in Figure S15 do not show enrichment of membranes.

Cell extracts obtained by chemical lysis are problematic because they contain surfactants. This comparison might not be meaningful.

The reviewer is correct about surfactants; However, this is only for illustration to show the crowd density of the cell extracts compared to live cells.

Unclear why "Bleach size" plots are shown. They are not discussed in the main text.

We show now a bleach size plot in Figure 2, where we explain the method. We removed them from the other figures.

Some figure panels have a strange aspect ratio, causing text to look distorted.

We corrected the figure distortion in the revised manuscript.

How are the values of D_confocal in buffer compared with past literature? Should these not all be diffusion limited? BCECF - larger than many of the drugs used here - shows ~ 100 μm^2/s in buffer (Verkman TiBS 2002).

We discussed this in our previous work (Ref. 13, iscience 2022, Dey et al.) Dconfocal is a relative diffusion rate and should not be confused with single-molecule diffusion coefficients. FRAP cannot measure the diffusion of more than 100 μm^2/s in the buffer. However, when comparing apparent FRAP rates between different fluorophores, it is not quantitative due to the major implication of the bleach radius towards diffusion rates. The rate constant normalized by bleach radius^2 is the proper way to compare i.e., our Dconfocal. (Ref. JMB 2021, iScience 2022 by Dey et al.).

Reviewer #2 (Recommendations For The Authors):

Recommendations:

(1) Page 3 at the bottom of the Introduction states, "...sodium azide (Hiruma et al., 2007) inhibited accumulation in lysosomes, cellular diffusion...increased only slightly." However, Figure 7C, F shows a sodium azide-induced decrease in the Dconfocal cellular diffusion. Please clarify.

Thank you for pointing this out; we corrected it in the revised version, including adding statistics.

(2) Page 6 states, "Quinacrine accumulation in the lysosome was observed also immediately after micro-injection, with aggregation increasing over time. Dconfocal of 4.2{plus minus}0.2 µm2 s-1 was calculated from line-FRAP immediately after micro-injection, slowing to 2.2{plus minus}0.1 µm2 s-1 following 2 hours incubations, with fractional recoveries of 0.63 and 0.57 respectively." If lysosome sequestration does not have an effect on cytosolic diffusion rates as the manuscript concludes, why do the authors think the diffusion rate decreased here within 2 hours? A solid conclusion would strengthen the conclusions of this manuscript rather than passing over it.

Thank you for pointing this out. We added the following text to page 7: “It is notable that the Dconfocal for Quinacrine remained consistent regardless of Bafilomycin treatment, 2 hours after incubation (Fig. S9D, 2.4±0.1 µm2s-1). However, when measured immediately after injection, the diffusion coefficient was higher at 4.2 µm2s-1 (Fig. S5D). This result does not support the notion that the faster diffusion measured immediately after cellular injection relates to lysosomal aggregation, and would better support self-aggregation, or aggregation with other molecules in the cell, which increases over time. This notion is further supported by the almost complete lack in FRAP observed 24 hours after injection (Fig. S5C).”

(3) In the Results section, the subheading states, "Inhibition of lysosomal sequestration is only slightly increasing diffusion in cells", but the conclusion for bafilomycin was...Dconfocal values were not altered by Bafilomycin A1", and the conclusion for sodium azide was diffusion coefficients (Figure 7B-C and 7E-F) were not much changed for the two drugs and stayed low... similarly to what was observed with Bafilomycin." The clear question is what is the result, "slightly increased diffusion, decreased diffusion, or had no significant effect at all"? Please clarify the wording in the manuscript to accurately describe the results.

Indeed, a small difference is obsevered between the two treatments. We added now statistical significance to Fig. 7D and H and to Fig. S8 and S9. In addition, we clarified this point in the text in p.7-8: “Comparative FRAP profiles and diffusion coefficients (Figure 7B-D and 7F-H) were slow, but conversely to Bafilomycin, sodium azide treatment did cause a further reduction is rates from Dconfocal 2.4±0.1 µm2s-1 to 1.8±0.1µm2s-1 for quinacrine and from 0.6 to 0.45 µm2s-1 for the GSK3 inhibitor (Figure 7C and G). Both Bafilomycin and sodium azide treatments resulted in elimination of drug confinement in the lysosome, and the small difference in diffusion rates may be a result of the de-acidification of the lysosomes by sodium azide, which may increase the protons in the cytosol upon treatment.”

(4) In Figure 8B, why was the Dconfocal for AM-fluorescein with or without sodium azide not included here? Besides consistency, the results might demonstrate significance. Please elaborate on the occlusion of this data.

Fraction recovery after FRAP of AM-fluorescein was very low. Calculating Dconfocal rates with such low fraction recovery is meaningless, as in the time of measurement only a small fraction recovered. Therefore, we calculated Dconfocal only when fraction recovery was at least 0.5.

(5) Throughout the Results section, the ideas and experiments are of relevance, but the suggestions/conclusions at the end of each paragraph of this section seem lightly thought out. For example, as stated on Page 8, "...however, this did not contribute new information to the puzzle." For a chemistry paper, a chemical suggestion strengthens the manuscript.

We want to thank the reviewer for these suggestions. We now made new Figures 9, and S16 to compare multiple parameters. Figure 9C shows a clear relation between pKa and Dconfocal, but no relation was found between logP, MW or number of aromatic rings and Dconfocal. Fig. S16 also shows the relation between drug concentration and Dconfocal values. We revised the discussion section to giver more weith to these quantitative assessments. These data are now discussed in p. 9.

In conclusion, the manuscript's ideas are needed, but the conclusions drawn from the experiments need to be strengthened, more explanatory, and consistent with the main conclusion of the manuscript.

See answer to point 5.

Version 2

Abstract

Abstract

For drugs to be active they have to reach their targets. Within cells this requires crossing the cell membrane, and then free diffusion, distribution, and availability. Here, we explored the in-cell diffusion rates and distribution of a series of small molecular fluorescent drugs, in comparison to proteins, by microscopy and fluorescence recovery after photobleaching (FRAP). While all proteins diffused freely, we found a strong correlation between pKa and the intracellular diffusion and distribution of small molecule drugs. Weakly basic, small-molecule drugs displayed lower fractional recovery after photobleaching and 10-to-20-fold slower diffusion rates in cells than in aqueous solutions. As, more than half of pharmaceutical drugs are weakly basic, they, are protonated in the cell cytoplasm. Protonation, facilitates the formation of membrane impermeable ionic form of the weak base small molecules. This results in ion trapping, further reducing diffusion rates of weakly basic small molecule drugs under macromolecular crowding conditions where other nonspecific interactions become more relevant and dominant. Our imaging studies showed that acidic organelles, particularly the lysosome, captured these molecules. Surprisingly, blocking lysosomal import only slightly increased diffusion rates and fractional recovery. Conversely, blocking protonation by N-acetylated analogues, greatly enhanced their diffusion and fractional recovery after FRAP. Based on these results, N-acetylation of small molecule drugs may improve the intracellular availability and distribution of weakly basic, small molecule drugs within cells.

Introduction

Drug molecules alter many biochemical reaction pathways inside the cell by interacting with proteins, DNAs, RNAs, or others. By this, they influence one or more specific cellular activities (Ong et al., 2009; Schreiber, 2005; Stockwell, 2004). Due to their bioactivity, stability, rather simple chemical synthesis allowing for industrial bulk scale production and the easiness of administration, most drugs are small molecules (< 1 kDa) (Ngo and Garneau-Tsodikova, 2018; Veber et al., 2002). Over half of these drugs have a significant degree of hydrophobicity and are weakly basic at intracellular pH of 7.4 (Charifson and Walters, 2014; Lipinski et al., 1997). Hydrophobicity is required to cross the lipid bilayer of the cell membranes, and polar functional groups solubilize the drug in the cellular medium (Lipinski et al., 1997). The balance between pKa and logP (partition coefficient of small molecules as measured from their partition between octanol/water) is critical for successful drug design (Manallack, 2007; Meanwell, 2011). Lipinski’s “Rules of 5” suggests that logP should be between 1 and 6 to be a candidate for successful oral administration. Small molecule drugs can be acidic or basic depending on their pKa values (Lipinski et al., 1997). Drugs with pKa ∼ 0-7 are considered acidic, and pKa ∼ 8-14 are regarded as basic or alkaline. Basic drugs with pKa ˃8 are more common than their acidic counterparts (pKa ˂ 7, ∼60:40 ratios). In some cases, lowering their pKa to the range between 6 and 7, has proven beneficial (Charifson and Walters, 2014). For a long time, mechanisms such as passive diffusion (Di et al., 2012) and carrier-mediated drug delivery (Dobson and Kell, 2008) or a combination of both (Sugano et al., 2010) have been used to explain intracellular drug delivery. However, transport of a drug across the cellular membrane is only one part of a successful delivery. Intracellular diffusion and bio-distribution needs to be taken in account when assessing the drug mode of action.

In previous work, we studied the diffusion of fluorescent drug molecules in crowded environments in vitro (Dey et al., 2022). However, the cellular cytoplasm of eukaryotic cells is by far more complex (Model et al., 2021). Intracellular membranes, multiple organelles (Ellis, 2001; Fulton, 1982) and a large diversity of macromolecules, in terms of size, charge and hydrophobicity, render the cellular interior extremely complex and heterogeneous (Milo and Phillips, 2015; Neurohr and Amon, 2020). Moreover, the pH inside cellular organelles can vastly differ from the cytosolic pH (Madshus, 1988). Organelles like lysosomes, late endosomes, Golgi, and secretory vesicles are acidic and have been known for sequestering weakly basic drugs (Asokan and Cho, 2002; Proksch, 2018). Therefore, predicting diffusion of a particular drug molecule in the dense cytoplasm is a complex task and requires sub-organelle resolution. For example, positively charged (basic) drugs designed to target nuclear DNA may bind cytoplasmic matter through nonspecific interactions (Long et al., 2022; Tiwari et al., 2018), they could be sequestrated within the lysosome, or diffuse slower than expected due to crowding and nonspecific interactions (Gotink et al., 2011; Jansen et al., 1999; Zhitomirsky and Assaraf, 2015).

The mobility of a drug in the cellular interior has been shown to be a good proxy for intracellular drug delivery and other processes, such as nonspecific binding or aggregation (Dey et al., 2022). The primary tool for measuring small molecule diffusion inside the cell is fluorescence recovery after photobleaching (FRAP), which inherently requires fluorescent drug molecules. FRAP with a confocal microscope is a sufficiently sensitive, versatile, and easy-to-use method suitable for measuring the diffusion of weakly fluorescent molecules. Unfortunately, single-molecule fluorescence correlation spectroscopy (FCS), a gold-standard method for measuring diffusion of fluorescent particles, cannot be applied for most small molecule drugs due to their weak fluorescence (Dey et al., 2022, 2021; Zotter et al., 2017).

In this work, we determined the intracellular diffusion coefficients of a series of fluorescent dyes and drugs compared to protein diffusion by FRAP (Figure 1 and Table 1). Specifically, we used Line-FRAP, which has a much higher time resolution and therefore can measure faster diffusion rates in comparison to standard FRAP (800 Hz versus 50 Hz, for detailed explanations, see (Dey et al., 2022) (Dey et al., 2021)). A pictorial representation of Line FRAP performed in a live HeLa cell treated with a fluorescent drug is shown in Figure 2A. The small red circle denotes the bleached area which is done by using a Clip tornado (of 2x2 pixel size). The white line indicates the scanning region which was performed along at a rate of 800 Hz in order to record the FRAP as a function of time. The Line-FRAP cross section profile as a function of time is shown in Figure 2B. The vertical direction of the line denotes the time progress. The vertically marked white rectangular Region of Interest (ROI), in the left panel (Figure 2B) shows the fluorescence intensity recovery over time, which is plotted in Figure 2C. The horizontal rectangular region of interest (ROI) outlined in white (in the right panel of Figure 2B) denotes the time frame immediately after the photo-bleach pulse (the highly intense green is the bleach pulse). The time interval between each line scanning measurement is ∼1.25 milliseconds. This horizontal white ROI generates the fluorescence intensity over the scanning line distance which is plotted in Figure 2D. This profile gives the effective bleach radius when fitted with a gaussian function (Dey et al., 2022, 2021). The apparent diffusion coefficients derived from FRAP measurements are termed Dconfocal are calculated using equation 1:

Chemical structures and pKa values (prediction from ChemAxon software) of the small molecules and drugs used in this study.

Schematic representation of Line-FRAP & Diffusion coefficients of proteins in the HeLa cells.

(A) Line-FRAP performed in a fluorescent drug labeled HeLa cell. The white line represents the scanning line; The red circular area denotes the bleaching area. (B): A single scanning line as a function of time (in the vertical direction) including the bleach is shown. The white vertical rectangular region of interest (ROI), which was used for analysis. A close-up view of the Line-FRAP profile, is also shown where the horizontal rectangular area (white colored marks ROI) denotes the fluorescence intensity as a function of length across the scanning line. (C) Average of 30 FRAP recovery curves as a function of time (N=30; Correlation coefficient of exponential fit: R=0.98) and (D) Averaged bleach size profile with gaussian fits (N=30; R=0.99) are shown. (E) Comparison of diffusion coefficients (Dconfocal in blue circles) and percentage of recoveries (in red squares) of bacterial proteins and BSA as measured in HeLa cell cytoplasm are shown. Error bars represent SE calculated from fitting the FRAP progression curves, which are averaged over at least 30 independent measurements.

where t1/2 is the half-time of recovery and re and rn are the effective and nominal bleach radii values (for details see (Dey et al., 2021)). Dconfocal values for proteins determined by Line-FRAP agreed with the diffusion rates measured by FCS (Dey et al., 2021). For small molecules, Line-FRAP only provides relative 3D diffusion coefficients (Dey et al., 2022). We studied the following fluorescent drugs: the antimalarial drugs quinacrine dihydrochloride (Ehsanian et al., 2011), primaquine phosphate (Camarda et al., 2019) and amidoquine (Olliaro and Mussano, 2009). The anti-cancer drug mitoxantrone (Shenkenberg and Von Hoff, 1986), and the glycogen synthase kinase GSK3 inhibitor SB216763 (Wagman et al., 2005). In addition, we monitored the diffusion of the diagnostic staining agent fluorescein disodium salt (Jampol and Cunha-Vaz, 1984), the fluorogenic substrate CCF2 (Zuverink and Barbieri, 2015) and boron dipyrromethene (BODIPY) analogues, which are used as fluorescent markers for lipids, membranes, and other lipophilic compounds (Poryvai et al., 2022).

The diffusion coefficients and their cellular distributions showed very slow diffusion of the basic compounds, i.e., low FRAP recovery, combined with an accumulation of the molecules in lysosomes. Surprisingly, while inhibiting the V-ATPase H+ pump using Bafilomycin A1 (Spugnini et al., 2010; Wang et al., 2021) or sodium azide (Hiruma et al., 2007) inhibited accumulation in lysosomes, cellular diffusion and fractional recovery increased only slightly. Contrary, blocking protonatable amino groups by acetylation greatly enhanced diffusion and FRAP recovery of studied small molecules. As over half the small molecule drugs are basic, our findings should be considered in drug assessment and development, as slow diffusion and low FRAP recovery is directly related to reduced cellular activity. Contrary to small molecules, we show here that diffusion rates within the cytoplasm of HeLa cells of 16 different E.coli proteins with pKa values of 4.5-8, are within the expected range, suggesting that hindered diffusion is mainly a problem of small molecules.

Results

Acidic Proteins diffuse freely in the HeLa cytoplasm

We have previously shown that diffusion in the cellular cytoplasm of E-fts and baeR, two bacterial globular proteins, aligns with the expected values considering the ∼2-3-fold higher macro-viscosity (Dey et al., 2021). Here, we extended this study to include 16 additional proteins (mostly from E. coli) whose oligomeric states in solution were previously characterized (Table S1 and (Marciano et al., 2022)). The aim was to determine how different heterologous proteins (E.coli proteins in HeLa cells) diffuse. The proteins showed Dconfocal values of 18-30 µm2s-1 in HeLa cell cytoplasm, which is typical for a cytoplasmic protein diffusion coefficient (Figure 2E). Moreover, full FRAP recovery was observed, indicating free diffusion. Table S1 provides the predicted isoelectric points and net charges at pH 7.4 of the purified proteins as calculated by Prot Pi, showing pKa values of 4.5-8, which are representative of the majority of E.coli cytosolic proteins. The results show that acidic and neutral proteins, even if not in their native environment, do not stick to the HeLa cell cytoplasm or membranes.

Negatively charged small molecules rapidly diffuse in live cells

Fluorescein is a negatively charged small organic molecule (pKa’s = 2.85, 3.7, M.W. = 332 Da, Figure 1 and Table 1). We showed in previous work that fluorescein diffusion is slower in the presence of crowder proteins like BSA, lysozyme and, myoglobin, even at relatively low protein concentrations (Dey et al., 2022). Here, the Fluorescein dye was micro-injected in live HeLa cells, followed by FRAP measurements (Figure 3 A-D). We found that the diffusion coefficient of fluorescein in PBS (Dconfocal of 56.5±2.4 µm2s-1) was reduced to Dconfocal of 38.5±2.0 µm2s-1 in HeLa cytoplasm (Figure 3C). In the cell nucleus, a value of Dconfocal of 43.2±2.2 µm2s-1 was determined, while in concentrated HeLa cell extract (equivalent protein concentration of ∼100 mg/mL), Dconfocal of 49.6±1.9 µm2s-1 was measured. The faster Dconfocal in cell extract compared to live-cell cytoplasm may result from the higher crowding density of the latter. The lowering of diffusion coefficients of negatively charged fluorescein correlates with the increasing crowding density of the medium moving from simple buffer to cells. For comparison, we also incubated HeLa cells for 24 hrs with fluorescein, which is sufficient time for it to accumulate in the cells by diffusion (Figure S1 A-B). While micrographs did not show dye aggregation inside the cell cytoplasm (Figure S1A-B) and diffusion rates are similar to those determined by microinjection (Figure S1C and S1D), the fractional recovery dropped from 0.95 to 0.65. This suggests that a fraction of fluorescein is immobile in the cell after 24 h incubation, and is not diffusing.

Fluorescein, CCF2 and CF514 diffusion in PBS and inside HeLa cells.

Comparative merged FRAP profiles (N= 30; R= 0.99 for each of the fits) in PBS and inside HeLa cells with exponential fits are shown for (A) Fluorescein, (E) CCF2 and (H) CF514. Comparative Dconfocal values for (C) Fluorescein (F) CCF2 and (I) CF514 are also shown. (B) Dconfocal values as calculated from individual FRAP curves for Fluorescein (see also Fig. S2). Dconfocal values as calculated from the merged FRAP curves shown in panel A. (D) HeLa cells after Micro-injection of Fluorescein; (G) CCF2; and (J) for CF514. Error bars represent SE calculated from fitting the FRAP progression curves of 30 merged independent measurements. Statistical significance calculations are detailed in the materials and methods section.

For calculating the Dconfocal values shown in Figure 3C, 30 individual Line-FRAP curves were merged for signal to noise reduction (Figure 3A) (Dey et al., 2022, 2021). For comparison, we calculated Dconfocal values from individual Line FRAP progression curves in HeLa cytoplasm (Figure S2A and S2B), obtaining the same mean value as for the merged curve (Figure S2B and S2D), however, with much larger variation over the mean. To estimate whether the large variation from the mean of the individual curves is due to different fluorescein concentrations in the individual measurements, we plotted the fluorescence signal (which is directly related to the concentration) versus the FRAP recovery rate (Figure S3A) and versus the percent recovery (Figure S3E). The lack of correlation suggests that the variation in FRAP rates of the individual measurements is due to the low signal to noise of the individual curves versus the average. Next, we calculated Dconfocal also for fluorescein in PBS, nucleus and HeLa cell extract (Figure 3B). The mean values from individual curves are very similar to those calculated from the merged curves (Figure 3C), with a normal distribution of the individual calculated Dconfocal values around the mean. Moreover, also for the values measured in the nucleus, no relation between fluorescein concentration and FRAP rates or percent recovery was observed (Figure S3B and F). Overall, this suggest that merged curves can be used to calculate Dconfocal values, as they give much better signal to noise, with apparent minimal loss of information.

CCF2 is a fluorogenic substrate used for monitoring β-lactamase activity, with a pKa of ∼5.1 (Zlokarnik et al., 1998). It is composed of cephalosporin core linking B7-hydroxycoumarin to fluorescein (Figure 1). Its calculated Dconfocal values in PBS and inside cells are about half as fast as those measured for fluorescein (Figure 3E-G), and its fractional recovery is ∼0.7. No sign of aggregation is seen in micrographs inside the cell cytoplasm or nucleus (Figure 3G). The substantial slower rate of recovery, and lower fractional recovery of CCF2 relative to Fluorescein suggests its diffusion to be obstructed within the cell. This is in line with our prediction, that the observed substrate limited enzymatic degradation of CCF2 in HeLa cells is a consequence of its occlusion within the cellular cytoplasm (Zotter et al., 2017).

A third molecule for which we determined the diffusion coefficient is the labelling dye CF514. It is used to label proteins and has high quantum yield. As its structure has not been published, we estimated the pKa value from the pH dependent UV-spectrophotometry measurements. The sharp changes of the UV-spectrum at pH∼3 (Figure S3I-J), as well as the redshift confirms a pKa value of ∼ 3. The FRAP curves look similar to fluorescein (Figure 3H-J), with ∼90% recovery after FRAP. Dconfocal in PBS was ∼ 61.1±3.0 µm2s-1, and 26.3±1.9 and 29.2±2.0 µm2s-1 in the cell cytoplasm and nucleus of HeLa cells, respectively (Figure 3I). The drop of Dconfocal is attributed to cellular crowding. The results so far show that acidic small molecules diffuse well inside cells. Next, we focus on basic small molecules with higher pKa values.

High FRAP recovery but slow diffusion measured for the glycogen synthase kinase (GSK3) inhibitor SB216763

The GSK3 inhibitor SB216763 (pKa=7.0, M.W.=371 Da, Figure 1) aggregates when dissolved in PBS buffer solution. However, in our previous study we have seen that aggregation of this particular small molecule drug can be reversed by adding BSA and the extent of de-aggregation depends on the amount of BSA added (Dey et al., 2022). Here, our aim is to compare diffusion behavior of the GSK3 inhibitor in solution phase to that found in live cell cytoplasm. Since the molecule has some inherent solubility issues in PBS buffer, we compared its diffusion also in BSA protein solution and in cell culture media DMEM (Dulbecco’s modified eagle medium). Comparative FRAP profiles and Dconfocal rates of SB216763 in different mediums are shown in Figure 4A-C. The FRAP profiles and relative Dconfocal values show very slow diffusion of SB216763 in HeLa cell cytoplasm of ∼0.61±0.03 µm2s-1 with a fraction recovery of 0.8 after 10s (Figure 4A-C). This is a 40-fold reduction in Dconfocal compared to measurements in 100 mg/mL BSA protein solution, where SB216763 is not aggregating (Dconfocal∼24.3±0.5 µm2s-1 and fractional recovery of ∼0.9 - Figure 4A-C). In cell culture DMEM media (containing 5% serum protein) fractional recovery after FRAP is very low (Figure 4A), indicating that the drug is still aggregating (Dey et al., 2022). Interestingly, the Dconfocal value in HeLa cell extract was higher ∼ 11.3±0.7 µm2s-1 as is its fractional recovery (0.9), showing that HeLa cell extract, like BSA is disaggregating SB216763. Co-localization with Lysotracker-633 (Figure 4D-E) shows that SB216763 colocalizes and is sequestered in acidic chambers of the cell lysosomes. Moreover, SB216763 accumulation into the lysosome increased over time (Figure 4D and E, 45-minute and 2-hour treatment). Varying drug dosages resulted in Dconfocal values of 0.42-0.61 µm2s-1, suggesting that the slow diffusion is not related to SB216763 concentration (Figure S4A-B). To validate Dconfocal values determined from merged progression curves, we calculated them also using individual Line FRAP profiles (Figure 4C vs 4B), with both methods showing the same mean values. Individual FRAP progression curves are shown in Figure S2E, which have a much lower signal to noise ratio in HeLa cell cytoplasm, but result in a similar mean Dconfocal value as determined from the merged curves (Figure 4A-C and S2E-H). There may be a slight decrease in FRAP rates and percent recovery at higher SB216763 concentrations as determined from the individual profiles (Figure S3C and G).

Diffusion of GSK3 inhibitor in PBS and inside HeLa cells.

Comparative (A) averaged FRAP recovery profiles with exponential fits for GSK3 inhibitor (SB216763) (N= 30; R= 0.99 for each of the fits). Dots show prebleach fluorescence, Black line for DMEM, blue for PBS buffer + 50 mg/mL of BSA, green for HeLa cells and orange for HeLa extract. (B) Dconfocal values calculated from individual FRAP curves. (C) Dconfocal values calculated from merger of 30 independent FRAP curves. (D-E) Colocalization of GSK3 inhibitor in lysosomal compartments. HeLa cells were treated with Lyso Tracker 633 for 30 mins. GSK3 inhibitor treatment for (D) 45 mins and (E) 2 hours respectively at 10 µM concentrations are also shown. Error bars represent SE calculated from fitting the FRAP progression curves, which are averaged over at least 30 independent measurements. Statistical significance calculations are detailed in the materials and methods section.

Despite its high solubility in aqueous solution, Quinacrine diffusion in HeLa cells is slow

Quinacrine DHC (M.W.= 472.9 Da, Figure 1) is a compound highly soluble in PBS (Dey et al., 2022) with a fast Dconfocal rate of 41.9±3.0 µm2s-1 (Figure 5A-C). This rate dropped to Dconfocal∼2.5±0.1 µm2s-1 in HeLa cell cytoplasm, with fraction recovery of ∼0.8, in comparison to one in buffer. In HeLa cell extract a Dconfocal rate of 15.1±0.6 µm2s-1 was determined, which lays between PBS and HeLa cytoplasm. Dconfocal estimated from individual Line FRAP profiles and the merged Line FRAP profile (Figure 5B vs 5C) are also shown. Figure S2I-L shows that the merged FRAP profile yields less noisy data and gives a better fit, yet the mean Dconfocal values are the same. As with SB216763, lower fraction recovery is observed at high quinacrine concentrations, but the FRAP rates seem to be concentration independent (Figure S3D and H). Micrographs of live HeLa cells show that the quinacrine DHC molecules aggregate in the cytoplasm. Co-localization with Lysotracker-633 (Figure 5D-E) show that quinacrine colocalizes and is sequestered in acidic chambers of the cell lysosomes. Moreover, quinacrine aggregation into the lysosome increased over time (Figure 5D and E, 45-minute and 2 hours treatment). Therefore, we measured diffusion instantaneously following micro-injection, and compared it to 2 and 24 hours of incubation (Figure S5A-B). Quinacrine accumulation in the lysosome was observed also immediately after micro-injection, with aggregation increasing over time. Dconfocal of 4.2±0.2 µm2s-1 was calculated from line-FRAP immediately after micro-injection, slowing to 2.2±0.1 µm2s-1 following 2 hours incubations, with fractional recoveries of 0.63 and 0.57 respectively. After 24 hours of incubation the fractional recovery is less than 0.25, and thus the diffusion value is meaningless (Figure S5C and D). Next, we compared FRAP of the cells treated with 2 and 6 µM quinacrine solutions after 2- and 24-hours incubations (Figure S6A-F). The concentration of quinacrine had only a marginal effect on the results. Again, following 24 hours of incubation, most of the quinacrine is in the lysosomes.

Diffusion of Quinacrine in PBS and in HeLa cells.

Comparative (A) averaged FRAP recovery profiles with exponential fits for Quinacrine dihydrochloride (N= 30; R= 0.99 for each of the fits), (B) Dconfocal values calculated from individual FRAP curves. (C) Dconfocal values calculated from merger of 30 independent FRAP curves measured in PBS buffer, 50mg/mL of BSA, HeLa cell extract and in live HeLa cells. (D-E) Colocalization of Quinacrine DHC in lysosomal compartments are shown. HeLa cells were treated with Lyso Tracker 633 for 30 mins. Quinacrine DHC treatment was done for (D) 45 mins and (E) 2 hours respectively at 10 µM concentration. Error bars represent SE calculated from fitting the FRAP progression curves, which are averaged over at least 30 independent measurements. Statistical significance calculations are detailed in the materials and methods section.

Mitoxantrone, primaquine and amidoquine all aggregate within cells

Mitoxantrone (anti-cancer) and primaquine phosphate (antimalarial) fluorescence with low quantum yields and weak photo-stability. However, due to their slow diffusion, the temporal resolution of standard XY-FRAP was sufficient for FRAP measurements (Figure 6A-H). Figures 6A and 6E show very low fraction recovery for these two drugs, suggesting that these weakly basic drugs are also sequestered in the cellular cytoplasm, as indeed shown in Figures 6B and 6F. Colocalization with LysoTracker red-633 show that Primaquine (S7A-B) and Amidoquine (S7C-D) are sequestered in the lysosomes. For amidoquine unfortunately, we could not collect any FRAP profiles as the dye did not survive the photobleached pulse. While we selected bleach spots to be small and located outside of lysosomes, this does not assure that some of the bleached area does not include smaller lysosomes. Therefore we investigated whether inhibiting lysosomal trapping will eliminate slow diffusion of cationic drugs.

Mitoxantrone and Primaquine diffusion inside HeLa cells.

Averaged XY-FRAP recovery profiles with exponential fits (N= 20; R= 0.98 for each of the fits) for (A) Mitoxantrone and (E) Primaquine in HeLa cells. Fluorescent and Transmission channel images of HeLa cells after 45 mins incubations with (B-C) Mitoxantrone and (F-G) Primaquine are shown. Time lapse micrographs of a portion of HeLa cell going through a classical rectangular XY-FRAP protocol for (D) Mitoxatrone and, (H) Primaquine are also shown.

Inhibition of lysosomal sequestration is only slightly increasing diffusion in cells

Lysosome internal pH of ∼4.5 is maintained by vacuolar proton ATPase (V-H+-ATPase) (de Duve et al., 1974; Kaufmann and Krise, 2007). The low pKa of weakly basic small molecule drugs drive their accumulation in the lysosome. There, they become protonated (ionic form is less-permeable), which hinders their back-diffusion to the cytosol. The phenomenon is known as cationic ion trapping (Asokan and Cho, 2002; Kaufmann and Krise, 2007) and results in drug accumulation in lysosomes up to millimolar concentrations (de Duve et al., 1974). We blocked lysosomal acidification using either Bafilomycin A1 or sodium azide prior to drug administration. Bafilomycin A1 is a specific V-ATPase inhibitor and thus inhibits the supply of protons to the lysosomal chamber (Wang et al., 2021). Sodium azide, blocks ATP generation, resulting in the disruption of cellular activities which require ATP (Hiruma et al., 2007; Ishii et al., 2014). As a result, cationic ion trapping is inhibited. These two compounds were used to investigate whether lysosomal trapping is the cause for the 20-40-fold lower diffusion and lower fraction recovery in HeLa cells compared to buffer. Figures S8A and B, and Figure S9A and B show colocalization images for GSK3 inhibitor and quinacrine, respectively (with Lysotracker-633 used as marker) after addition of 100 nM Bafilomycin A1 (Yoshimori et al., 1991). The micrographs show that Bafilomycin A1 strongly reduces lysosomal volumes following quinacrine and GSK3 treatment (Figure S8A vs S8B and S9A vs S9B). However, Dconfocal values were not altered much by Bafilomycin A1, as shown in Figures S8C-D and S9C-D for GSK3 inhibitor and Quinacrine, respectively. Dconfocal for the GSK3 inhibitor is ∼ 0.60±0.02 µm2s-1, and for Quinacrine ∼ 2.4±0.1 µm2s-1. This suggests that Lysosomal accumulation of these two drugs is not the reason for the slow diffusion measured in the HeLa cell cytoplasm. It is notable that the Dconfocal for Quinacrine remained consistent regardless of Bafilomycin treatment, 2 hours after incubation (Fig. S9D, 2.4±0.1 µm2s-1). However, when measured immediately after injection, the diffusion coefficient was higher at 4.2±0.2 µm2s-1 (Fig. S5D). This result does not support the notion that the faster diffusion measured immediately after cellular injection relates to lysosomal aggregation, and would better support self-aggregation, or aggregation with other molecules in the cell, which increases over time. This notion is further supported by the almost complete lack in FRAP observed 24 hours after injection (Fig. S5C).

To verify the Bafilomycin results, HeLa cells were pre-treated with Sodium azide at 1 and 50 mM concentrations (Figure 7A and E), followed by treatment with GSK3 inhibitor or Quinacrine (Poste and Papahadjopoulos, 1976). The micrographs in Figures 7A and 7E, show that sodium azide inhibited lysosomal accumulation of these drugs. Comparative FRAP profiles and diffusion coefficients (Figure 7B-D and 7F-H) were slow, but conversely to Bafilomycin, sodium azide treatment did cause a further reduction is rates from Dconfocal 2.4±0.1 µm2s-1 to 1.8±0.1µm2s-1 for quinacrine and from 0.6 to 0.45 µm2s-1 for the GSK3 inhibitor (Figure 7G and C). Both Bafilomycin and sodium azide treatments resulted in elimination of drug confinement in the lysosome, and the small difference in diffusion rates may be a result of the de-acidification of the lysosomes by sodium azide, which may increase the protons in the cytosol upon treatment.

Effect of Sodium Azide on the diffusion of GSK3 inhibitor and Quinacrine inside HeLa cells.

(A) GSK3 inhibitor and (E) Quinacrine dihydrochloride treated HeLa cells with or without sodium azide. Comparison of merged FRAP profiles with exponential fits (N= 30; R= 0.99 for each of the fits) and calculated Dconfocal values from individual fits and combined averaged fit for (B-D) GSK3 inhibitor and (F-H) Quinacrine dihydrochloride are shown. Dconfocal estimations from individual FRAP curves are shown in (C,G). Error bars represent SE calculated from fitting the FRAP progression curves, which are averaged over at least 30 independent measurements. Statistical significance calculations are detailed in the materials and methods section.

5-amino fluorescein diffuses slowly in HeLa cells and is accumulated within lysosomes

To experimentally test the relation between the pKa of small molecules and their in-cell diffusion, we measured the diffusion of 5-amino fluorescein (AM-Fluorescein, M.W.= 347.3 Da, Figure 8A), which is structurally similar to fluorescein disodium salt, but with an additional -NH2 group that increases its pKa from ∼3 to ∼9 (Figure 8A). Figures 8B and C show comparative FRAP traces for these molecules, and for CCF2. While fluorescein disodium salt has a fast FRAP rate, and is completely recovering, the cationic charged AM-fluorescein recovers much slower with a fractional recovery of 0.4. Moreover, while fluorescein disodium salt is not sequestered in lysosomes, 5-Amino fluorescein is. This sequestration is alleviated by adding sodium azide to the cells (Figure 8D) but has only a small effect on the FRAP rate and fractional recovery (Figure 8B). This experiment strongly suggests that the functional primary amine group is the main contributor towards the slow diffusion and low FRAP recovery of AM-fluorescein in the cell cytoplasm.

FRAP of Fluorescein analogues inside HeLa cells.

(A) Structures of Fluorescein analogues with pKa values. (B) FRAP recoveries with exponential fits for Fluorescein, CCF2 and 5-amino Fluorescein, with/without Na-Azide treatments in HeLa cells (N= 30; R= 0.99 for each of the fits) . (C) Line FRAP profiles with time lapses. Micrograph images of treated HeLa cells after small molecule incubations.

Acidification of primaquine and BODIPY increases their diffusion and FRAP recovery

We substituted the NH2 of primaquine (antimalarial drug) and BODIPY compounds for the charge-neutral-NHAc functional group (Figure S10A and S11A). Due to their low quantum yield, we could not implement the faster Line-FRAP protocol, therefore, we used standard FRAP, which suits the slow diffusion of both compounds. As seen from the FRAP profiles (time-dependent monographs) in Figures S10B-D for primaquine derivatives and in Figures S11B-C for BODIPY derivatives, the recovery is much slower for free amine-containing moieties. Figure S10I-J shows that the fractional recovery for primaquine is only ∼0.27, whereas, for primaquine-NHAc, the recovery is ∼0.6. Comparing the measurements using micro-injecting versus incubation of the cells for FRAP measurements for primaquine-NHAc gave similar results (Figure S10E-H). Unfortunately, the classical FRAP is not fast enough to calculate the diffusion coefficients, as the dead time was 150 ms. For the BODIPY analogues, the observations were similar, but to a smaller extend. For BODIPY-NH2, the recovery percentage is ∼57%, whereas, for BODIPY-NHAc, the recovery is ∼67% (Figure S11H). The estimated recovery half-life also supports the faster diffusion of BODIPY-NHAc, as shown in Figure S11I.

Basic small molecules do not colocalize with lipid droplets, the ER or nucleic acids

To identify the location of colocalization of basic small molecules, we tested colocalization of GSK3 inhibitor in lipid droplets using the Nile red dye (Figure S12) and in the ER using a specific BFP/mCherry tagged fluorescent antibody markers (Figure S13). Next, we looked for colocalization of Mitoxantrone with nucleic acids present in cytoplasm using SYTO blue (Figure S14). No colocalization was found between the drugs and either lipid droplets, the ER or nucleic acids.

Super resolution images of GSK3 inhibitor treated HeLa cells were taken to improve the spatial distribution (Figure S15); which however, this did not contribute new information to the puzzle. Therefore, we could not pinpoint the location of sequestration of weakly basic small molecule drugs within the cellular cytoplasm, except the lysosome.

Discussion

In this study, we selected well-known and widely used small molecule drugs to study their behavior in aqueous solutions and living cells. All drug molecules used here are fluorescent, allowing for their FRAP measurements. Previously, we found that small-molecule drugs diffuse differently in crowded media than in a simple buffer solution, even if aggregation is not an issue (Dey et al., 2022). Here, we show that many small molecules get trapped within the cellular cytoplasm, resulting in extremely slow diffusion and lower fractional recovery after FRAP. This is particularly observed for cationic-charged small molecules, with Dconfocal values of 20-40-fold lower than observed for fluorescein or CF514 (Figure 9A vs 9B), which, in many cases is accompanied by low recovery after photobleaching. The later suggests that most molecules are occluded to components of the HeLa cell cytoplasm. Even dense cell extract from HeLa cells is not mimicking the in vivo results (Figure 4A-C and 5A-C). This is very different from protein diffusion, where none of the 16 proteins measured here showed slow diffusion or low fractional recovery, despite their E.coli origin, expressed in HeLa cells. Slow diffusion and lower fractional recovery were limited to basic small molecules, showing Dconfocal values of 0.2-2 µm2s-1. While the fractional recovery for all proteins is ∼0.9, for many basic small molecules it is 0.2-0.5.

Dconfocal of small molecules in PBS buffer and inside HeLa cells.

Comparison of Dconfocal values of small molecule drugs in (A) PBS buffer and in (B) HeLa cells. Error bars represent SE calculated from fitting the FRAP progression curves, which are averaged over at least 30 independent measurements. Dependency of Dconfocal values on (C) pKa and (D) logP values in HeLa cells. # in (A) denotes DMEM media (instead of PBS), * in (B) denotes measurements done in the nucleus.

As primary culprit for the slow diffusion of the basic small molecule drugs we suspected sequestration in the lysosomes. Indeed, high level of accumulation in the lysosomes were found for all the basic drugs. However, pre-treatment with sodium azide or Bafilomycin A1, which we showed to inhibit lysosomal accumulation, had no or only a small effect on diffusion or fractional recovery after FRAP, which stayed low. Therefore, we evaluated colocalization of GSK3 inhibitor with lipid droplets, the ER or nucleic acids present in cytoplasm (Figure S12-14), but did not detect any.

The conclusive evidence that slow diffusion and lower FRAP recovery results from the drugs being basic came by altering three molecules from being basic to acidic and vice versa. The best example here is fluorescein, where transformation to 5-amino fluorescein drastically reduced its diffusion as well as its fractional recovery after FRAP. Two additional examples here were Primaquine that was modified to Primaquine Acetate and BODIPY-NH2 that was modified to BODIPY-NHAc. In both cases, the diffusion became faster, and the fractional recovery increased significantly. We did not find any specific correlation between individual average diffusion rate constants with the concentration of small molecule drugs inside the cell (Figure S3A-D). For fraction recovery, we saw some decrease at high concentrations for the GSK3 inhibitor and quinacrine, but not fluorescein (Figure S3E-H), which would support out notion that they are sequestered in the cell. Correlating the calculated Dconfocal values with pKa shows a clear trend, where diffusion is strongly slowed down at higher pKa values (Figure 9C). Conversely, logP, molecular weights or number of aromatic rings present in the drug molecules did not relate to Dconfocal values (Figure 9D and S16A-B). This finding is important, as the activity of a drug within the cell is dictated by its active concentration. When the drug is sequestered, its active concentration is reduced. Even, if it is not fully sequestered, but only diffuses much slower, it will result in lower association rate constant, and thus lower affinity towards the drug target. As a result, to keep the drug active, its dose has to be increased, which can have negative implications on side effects, due to off-target binding.

Conclusions

Our study shows that in vitro biophysical crowding studies for small molecule drugs are of limited value if one wants to understand the biophysical behavior of the same drug within the cell. While it is true that availability of a drug within a cell is only one factor dictating its biological activity, it is a crucial one. Our findings here raise an important limitation on the standard rules for drug design, as these do not consider the stickiness of basic small molecules within the cell. We were able to directly address this question by using fluorescent molecule drugs, and measuring their location and diffusion within cells. Most importantly, we also show that by blocking protonation of a number of basic compounds we were able to increase substantially their diffusion and recovery after photobleaching, and thus potentially their activity. These findings may be consequential in future drug development.

Materials and methods

All the reagents used are described in a table format (Table S2, Key resources). The supplementary method sections describe the synthetic procedures, characterizations, and purity of the synthesized compounds.

Selection of drugs used in this study

To identify small molecule drugs that have fluorescence in the visible light, we screened in high-throughput 384 well plates 1600 commonly used, small molecule drug compounds for their penetration into HeLa cells and fluorescence at the DAPI (377/447), FITC (475/520), TexasRed (560/624) and Cy5 (631/692) channels. As positive control, we used Pyrvinium (red), Doxorubicin (green)and CSB (blue). Out of those, 16 compounds gave positive results. These compounds were then verified for their photostability, availability and solubility. This resulted in the list of compounds shown in Table 1, which includes also compounds used by us in previous studies.

Mammalian Cell Culture. HeLa cells were grown in 35-mm glass-bottomed dishes (MatTek Corporation) in DMEM (1X) Gibco (Life Technologies Limited) supplemented with 1X pyruvate, penicillin/streptomycin (BioIndustries), and 10% fetal bovine serum (Life Technologies Limited). The cells were subcultured when 80% confluence was reached using trypsin-EDTA for cell detachment. 2x105 HeLa cells in 2.5 ml of DMEM were pipetted into glass-bottomed dishes and incubated overnight. The cells were cultured in a humid atmosphere at 37°C and 5% CO2. The cells were imaged 24–30 h after seeding. Before the microinjection, the medium was aspirated, and fresh medium was supplemented with 25 mM HEPES, pH 7.4. For drug treatment, HeLa cells were incubated with 10 µM of drug (diluted 1:1000 from stock) for indicated times at 37 °C, followed by three times PBS 1X wash prior to imaging. For dose-dependent studies, the drug concentrations varied typically between 2-12 µM. Sodium azide concentration was 1 and 50 µM. Bafilomycin A1 was used at 100 nM concentration. Similarly, during the colocalization study, HeLa cells were treated with LysoTracker dye (1X) with 1:1000 dilution for 30 mins to 1 h at 37 °C, followed by 3x washing with PBS. Quinacrine, GSK3 inhibitor, Mitoxantrone, 5-amino Fluorescein, Amidoquine, Primaquine and BODIPY analogues were incubated with cells, while Fluorescein, Quinacrine, Primaquine-Ac, and CF514 dye were also micro-injected inside the HeLa cells.

Microinjection into HeLa cells

Microinjections were performed using the Eppendorf FemtoJet microinjector attached to the Eppendorf InjectMan NI2 micromanipulator. The fluorescein sodium salt, CCF2, CF514, quinacrine, primaquine analogues, BODIPY analogues were dissolved in DMSO and highly concentrated stock aliquots were made. Diluted PBS solutions of small molecules were injected into cells using glass capillaries from Warner instruments and pulled by a vertical puller (Narishige). For every measurement, a single pressure pulse was applied to deliver the sample into the cell. Air was administrated at 15– 25 hPa for 0.1-0.3 s. For injections, single cells containing morphologically healthy and well-connected HeLa cells were selected. Before and after the microinjection, cell morphology and membrane integrity were confirmed by visually inspecting the injected cells.

HeLa cell extract preparation

Cytoplasmic HeLa cell extracts were prepared as described previously with slight modifications (Zotter et al., 2017). HeLa cells from 4×10 cm plates at 80% confluency were washed, treated with trypsin and collected. Pellets were mixed with 500µl of RIPA buffer, IMP40, 5 µL of protease inhibitor and stored in ice. After 20 min, the mixture was centrifuged at 13k×g. This process was repeated twice to obtain concentrated extract. A BSA calibration curve determined the total protein concentration; the final concentration was up to 100 mg/mL. After liquid nitrogen freezing, the final cell extract solution was stored at -80° C.

Protein purification and dye labeling

Bacterial proteins used in this study were purified as described by us previously (Marciano et al., 2022). The dye labelling procedure with CF514 for purified proteins and removal of excess dye was described in detail (Dey et al., 2021).

Confocal microscopy and FRAP Analysis

Confocal microscopy. Images were collected with an Olympus IX81 FluoView FV1000 Spectral/SIM Scanner confocal laser-scanning microscope, using 60X DIC oil-immersion objective, N.A. 1.35. For fluorescein sodium salt fluorescence measurements, excitation was done at 440 nm, using a diode laser at an output power of 1-4 % of maximal intensity for high to low concentrations. In contrast, emission was recorded from 520 to 550 nm using the spectral detection system. For CCF2-FA, excitation was done with the laser at 440 nm using 1-2 % of the maximal intensity, while emission was collected from 470-560 nm with SDM560 emission dichromator as cut-off filter. For quinacrine DHC and GSK3 inhibitor, excitation was done at 488 nm laser using 1-2% of the maximal intensity, while emission was collected from 502-560 nm with SDM560 emission dichromator as cut-off filter. During colocalization experiments of quinacrine DHC and GSK3 inhibitor with LysoTracker-633 dye, the same setting was used for detecting the small molecule drugs in Channel 1 (green). LysoTracker-633 dye was detected in Channel 2 (red) by exciting at 635 nm laser using 1-2 % of the maximal intensity, while emission was collected from 655-755 nm. Colocalization for Mitoxantrone (red channel) was done by using quinacrine to track the Lysosomes (green channel). Percentage of volume and drug intensity (material) above threshold is defined as percent colocalized (Figure S16C-D). For mitoxantrone and BODIPY analogues, excitation was done at 635 nm laser using 1-2 % of the maximal power, and emission was collected from 655-755 nm in Channel 2 (red). In the blue channel, excitation was done by 400 nm laser using 2-4% of maximal power, whereas the emission was collected from 420-480 nm with SDM480 emission dichromator as cut-off filter. The same setting is used in the blue channel for the SYTO blue maker (to visualize free nucleic acids) and blue fluorescent protein marker (BFP to visualize the ER). For the mcherry protein marker (to visualize the ER), the same setting for the red channel (Channel 2) is used. For every colocalization experiments, proper blank experiments were performed separately in all the respective channels with the same microscope settings to avoid any bleeding/leakage of emission signals from one Channel to another Channel, as shown in Figure S17. Image analyses were performed using FluoView/Imaris software, and data analyses were performed using Kaleidagraph software version 4.1 (Synergy). Colocalization coefficients as a percentage of material/volume above threshold colocalized are calculated using Imaris software version 7.

Line-FRAP and Classical XY-FRAP

Line-FRAP was carried out in liquid drops and HeLa cells. For photobleaching, “Tornado” of 4 pixels (2x2, 1-pixel = 0.207 micrometer) as default laser settings by the Olympus Fluoview confocal system (Dey et al., 2021). It is the smallest area achievable using Tornado. The bleach circle was kept precisely in the middle of the scanning line. The lines we scanned were unidirectional with time intervals of 1.256 ms 1000 times (equivalent to 1.256 s) in the majority of the measurements. The number of scans prior to, during, and after the photobleaching was 10, 42, and 948, respectively. Only in some specific cellular measurements (for GSK3 inhibitor and Quinacrine DHC, where recovery is very slow) the unidirectional lines were scanned 5000 times. Photobleaching was achieved by using the laser at 405-nm or with 635-nm excitations for the 63-millisecond duration at full intensity (100%). The simultaneous scanner moved at a 100 µs/pixel speed to perform an efficient photo-bleach. We have used two simultaneous scanners during the Line-FRAP experiments: one scanner (at 405 nm with the full intensity of 100%) for photobleaching and another scanner (at 440/515/635 nm with weak intensity) for data acquisition. For all the drugs except Mitoxantrone and BODIPY analogues (emission in red wavelengths), photo-bleach was performed by 405 nm laser. For fluorescence signal detection: fluorescein 440 nm laser (1-4%); CCF2 440 nm laser (1-2%); GSK3 inhibitor 440 nm laser (0-2%); quinacrine (1-2%) of maximal intensity were used. Emission collections were done from 520-550 nm for quinacrine DHC and GSK3 inhibitor.

Line FRAP was not possible for primaquine analogues, as it suffers from strong photo-bleach during recovery. Here classical XY FRAP was performed, with photo-bleach done by 405 nm laser with full intensity. The frame rate is maintained at 172 ms with a bleach pulse duration was 150 ms. Due to microscope setting limitations, Line FRAP is also not feasible for Mitoxantrone and BODIPY analogues. A classical XY-FRAP using two simultaneous scanners is done for these molecules. Here, bleach was performed by simultaneous 635 nm laser with the full intensity of 100%, and the main excitation was done by 635 nm laser (with the power of 4%). Emission collections were done from 655-755 nm. For BODIPY analogues FRAP recoveries, the bleach pulse time was 150 ms. The bleach pulse time was also 150 ms for primaquine analogues for comparative FRAP profile collections. The frame rate is maintained at 172 ms. Using the Olympus IX81 FluoView FV1000 Spectral/SIM Scanner confocal laser-scanning microscope, using Tornado (which requires SIM scanner to be loaded) greatly enhances bleaching efficiency. In addition, it shortens the time to obtain the first measurement after bleach (which is immediate in this mode). This property is highly beneficial for Line-FRAP measurements, where the time scale of data acquisition plays an important role. The Fluoview SIM scanner unit synchronizes laser light simulation with confocal and multiphoton imaging to avoid interruption to image observation during laser stimulation or manipulation. We have varied the intensity of the lasers to achieve an optimal signal/noise ratio. Fluorescence recovery plots were fitted to a double exponent growth curve. FRAP experiments were also performed inside the PBS buffer drops and crowding conditions. Line FRAP was done on the cells taken from different plates, treated independently. A coverslip was introduced at the top of the drop to stop evaporation during measurements. Our previously developed method on Line FRAP protocol has been employed to calculate diffusion coefficients from the FRAP rates and averaged bleach sizes. Diffusion rates cannot be calculated using XY-FRAP. However, FRAP parameters were kept constant so that qualitatively we could compare the average half-life and percentage of recoveries. When the areas of bleach were selected in the drug-treated cell cytoplasm, we avoided the lysosomes as much as possible, within the resolution limits of the confocal microscope. Lysosomes themselves were measured to move within the cytoplasm with an diffusion coefficient of 0.03-0.071 µm2 s−1 (Bandyopadhyay et al., 2014), which is much slower than the diffusion measured for even the slowest compounds using fast Line FRAP, further validating that we did not measure lysosome diffusion.

Determination of Dconfocal from FRAP results

We previously developed (Dey et al., 2021) the FRAP in Line mode (Line FRAP) method to monitor the diffusion rates of proteins in various environments. Line-FRAP allows a much faster data acquisition rate compared to classical XY FRAP. It is more suitable for measuring diffusion rates for fast diffusing molecules. To calculate Dconfocal, 20-30 independent measurements on different cells (n) were binned, and the curve fit of the progression curve was used to obtain t1/2, and re values and their associated errors as determined from the fit of the curve. rn was obtained from bleaching a fixed sample. The Dconfocal values were then calculated using equation 1 (see introduction and (Dey et al., 2021)). The standard errors (SE) of the individual parameters were combined to obtain Dconfocal±SE. The errors and means were used to calculate the significance level of the difference between two treatments, with ns for non-significant, * being for 0.05-0.01, ** being for 0.01-0.005, *** for 0.005-0.001 and **** <0.0001. Dconfocal values from individual Line FRAP profiles were calculated the same way, and one-way ANOVA statistical tests were performed to calculate the significance between treatments using the standard function available in GraphPad Prism v10.1 using Dunnett for multiple comparisons. To verify SE values, we repeated the ∼20-30 measurements independently multiple times, which gave the same SE as obtained from individual curve fits of n = 30 cells.